---
title: "Extending specvec: weightings, factorizations, methods"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending specvec: weightings, factorizations, methods}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(specvec)
```

The methods that ship with SpecVec are points on a grid, and the grid is open.
A new weighting, a new factorization, or a new ordinal cover scale enters by
registration, and from that moment it behaves like a built-in. This vignette
adds three things that run: a van der Maarel cover scale, a Jaccard association
weighting with its method, and a deliberate capability mismatch that the engine
rejects. The companion `vignette("specvec-methods")` covers the built-in
operators those examples build on.

## The registry model

A method in SpecVec is a registered name that points at a
`(weighting, factorization)` pair together with a capability descriptor:
its input shape, its native output, whether it reads cover, and a default
dimension. `register_method()` writes that record; nothing is copied. Adding the
hundredth method costs the same as adding the second, because dispatch reads the
record rather than branching on a method name. The five shipped methods are
exactly five such records. `pmi` and `abund_pmi` reuse the `eigen` factorization
with different weightings, `ca` and `glove` pair their own operands with `svd`
and the GloVe objective, and a sixth method that reused an existing weighting
would be one more `register_method()` line rather than one more code path.

```{r methods}
specvec_methods()
```

Three registries sit behind those names. `register_weighting()` records the
function that turns a `specvec_data` object into the operator a factorization
consumes. `register_factorization()` records the function that turns that
operator into a species by dimension matrix. `register_method()` records the
pairing plus its capability flags. A weighting takes `(data, ks, n_plots,
min_cooccurrence)`, where `data$P` is the sparse plot by species presence
matrix, `data$COV` is cover or `NULL`, `data$species` holds the character ids,
and `ks` indexes the species columns that survived the rare-species filter. It
returns a list with a `kind` field.

That `kind` field is the contract between the two halves. A weighting declares
one of three kinds, and that declaration alone decides which factorizations can
consume it. `"sym"` is a symmetric species by species matrix `$M`, the form an
eigendecomposition wants. `"counts"` is a species by species co-occurrence
matrix `$M`, the raw operand the GloVe objective brings its own loss to.
`"implicit"` carries matrix-vector closures rather than a stored matrix, so a
large contingency table is never densified, which is how correspondence analysis
runs. Alongside `$M` (or the closures), a weighting also returns `species`, the
character ids of the kept species in the row and column order of the operator, so
the factorization can attach names to the rows of the matrix it produces. A
factorization declares, through its `accepts` field, exactly one kind it will
consume: `eigen` accepts `"sym"`, `svd` accepts `"implicit"`, `glove` accepts
`"counts"`. Pairing a weighting with a factorization that accepts a different
kind is the one error the fitter raises before doing any numerical work, and the
worked mismatch later in this vignette shows the message it produces.

The capability descriptor on a method carries a few more flags. `input` is
`"species_species"` when the operand is a square species matrix and
`"plot_species"` when it is the plot by species contingency that correspondence
analysis factorizes implicitly. `native_output` is `"species"` for a method that
produces species vectors directly, or `"species_and_community"` for one that
yields plot vectors in the same pass. `supports_abundance` records whether the
method reads cover, which lets a caller know an embedding will respond to the
`COV` matrix rather than only presence. `default_dim` is the dimension used when
none is given. These flags are descriptive: they let the print methods report what
a method does without running it, and they are how a method declares its shape
rather than discovering it mid-fit.

The dataset for the rest of the vignette is a small latent-niche simulation. Each
of 60 plots draws a niche center on the unit interval; a species enters a plot
with probability falling off as the squared distance from its own niche optimum,
and cover is a random percent. Species with similar optima end up co-occurring,
which is the structure an embedding should recover. The rare-species filter stays
low (`min_occurrence` of 2) because the simulation is small. Twelve species and
sixty plots keep every operator small enough to print a corner of, and the
embedding dimension stays at six or fewer, which is more than the latent structure
needs and small enough to fit instantly. `set.seed(1)` fixes the simulation so the
numbers below reproduce.

```{r data}
set.seed(1)
n_plots <- 60L; n_sp <- 12L
niche <- runif(n_sp)
rows <- lapply(seq_len(n_plots), function(p) {
  prob    <- exp(-((niche - runif(1))^2) / 0.02)
  present <- which(runif(n_sp) < prob)
  if (length(present) < 2L) present <- order(prob, decreasing = TRUE)[1:2]
  data.frame(plot = sprintf("p%02d", p),
             species = sprintf("sp%02d", present),
             cover = round(runif(length(present), 1, 100)))
})
df <- do.call(rbind, rows)
x  <- specvec(df, plot = "plot", species = "species", abundance = "cover")
x
```

## Inspecting operators with cooc_matrix()

Before writing a weighting it helps to see what one produces. `cooc_matrix()`
runs a built-in weighting and hands back the operator, so the species by species
form is visible without fitting anything. The counts operator is the plain
co-occurrence matrix: entry `[a, b]` is the number of plots holding both species,
and the diagonal is each species' occurrence count.

```{r cooc-counts}
Cc <- cooc_matrix(x, "counts", min_occurrence = 2)
class(Cc)
dim(Cc)
```

The result is a `dgCMatrix`, sparse and square over the kept species. Most
species pairs never share a plot in 60 records, so most entries are structural
zeros that the sparse format never stores. The PPMI operator runs the same
co-occurrence count through a positive pointwise mutual information transform:
the count is compared against the product of the two marginals, the logarithm of
the ratio is taken, and anything below zero is clipped to zero.

```{r cooc-ppmi}
Cp <- cooc_matrix(x, "ppmi", min_occurrence = 2)
dim(Cp)
as.matrix(Cp)[1:5, 1:5]
```

The dense corner shows the symmetry the eigensolver relies on: `[sp01, sp02]`
equals `[sp02, sp01]`, the diagonal is zero because a species carries no mutual
information with itself, and a pair that never co-occurs reads as zero. That
symmetric, non-negative, sparse matrix is precisely the `kind = "sym"` operator
a new weighting needs to produce if it wants to feed `eigen`. The counts matrix
above is the same idea one transform earlier, the `kind = "counts"` operand. With
those two shapes in hand, writing a weighting becomes filling in the transform
between them.

`cooc_matrix()` is also the place to confirm that `min_occurrence` and
`min_cooccurrence` do what an extension expects. `min_occurrence` is applied
before the operator is built: species seen in fewer plots than the cutoff are
dropped from `ks`, so they never reach the weighting at all, and both operators
above are square over the same surviving species. `min_cooccurrence` is the
weighting's own business; it arrives as the fourth argument, and a species by
species weighting decides whether to prune pairs that share too few plots. A new
weighting receives the same two controls every built-in does, which keeps the
rare-species behaviour consistent across methods rather than reimplemented in
each.

## Worked example A: a new cover scale

Vegetation surveys record abundance on ordinal scales, and the one a survey used
is rarely the three SpecVec ships with. The van der Maarel 1-9 ordinal scale maps
nine class codes to percent-cover midpoints. Registering it makes those codes
convertible the same way Braun-Blanquet codes already are. A cover-scale function
takes the raw column `x` and an optional `mapping`, and returns a numeric
proportion in `[0, 1]`, with `NA` for any value it does not recognize. The `NA`
is a signal the imputation step reads: unrecognized or missing entries are filled
downstream with the column median.

```{r vdm-register}
register_cover_scale("vdm", function(x, mapping = NULL) {
  m <- c(`1` = 0.1, `2` = 0.5, `3` = 1.5, `4` = 3, `5` = 8,
         `6` = 18, `7` = 37.5, `8` = 62.5, `9` = 87.5)
  key <- as.character(x)
  out <- unname(m[key])
  pmin(pmax(out, 0), 100) / 100
})
```

The body looks up each code in the named midpoint vector `m`, where class 1 is
0.1 percent cover and class 9 is 87.5 percent. `as.character(x)` coerces the
column to the lookup keys, `m[key]` returns the midpoints (with `NA` for any code
absent from `m`), `pmin(pmax(out, 0), 100)` clamps to the valid percent range,
and dividing by 100 returns the proportion. The scale is now usable through
`cover_from_scale()`, the front door every scale shares.

```{r vdm-use}
cover_from_scale(c("1", "5", "9"), "vdm")
```

Class 1 returns 0.001, class 5 returns 0.08, class 9 returns 0.875, matching the
midpoints divided by 100. To feed a survey recorded on this scale into `specvec()`,
convert the column first and pass the result as proportions. The `cover_scale`
argument of `specvec()` is restricted to the three built-in names, so a freshly
registered scale is applied at the `cover_from_scale()` step rather than passed by
name into `specvec()`.

```{r vdm-pipeline}
veg <- data.frame(plot = df$plot, species = df$species,
                  vdm = sample(as.character(1:9), nrow(df), replace = TRUE))
veg$cover_prop <- cover_from_scale(veg$vdm, "vdm")
xv <- specvec(veg, plot = "plot", species = "species",
              abundance = "cover_prop", cover_scale = "proportion")
xv$meta$has_abundance
```

The converted column carries proportions in `[0, 1]`, so `cover_scale =
"proportion"` is the right reader for it, and the resulting object holds cover the
same as any other. A custom scale registered once is then a one-line conversion
wherever that survey's codes appear.

The `mapping` argument is the reason a scale is a function rather than a fixed
table. Braun-Blanquet conventions vary by author, and the split classes (2a, 2b,
2m) are not universal, so the built-in scale accepts a `mapping` that overrides
its default midpoints. The van der Maarel scale above ignores `mapping` because
its nine classes are fixed, but a scale whose midpoints a study revises can read
the override and fall back to a default when it is `NULL`. Returning `NA` for an
unrecognized code keeps the conversion honest: a stray code becomes a missing
value the imputation step handles, so a typo in one record stays contained to
that record.

## Worked example B: a new weighting and method

A Jaccard association is a natural weighting that SpecVec does not ship: the
fraction of plots holding either of two species that hold both. It is symmetric
and bounded in `[0, 1]`, which makes it a `kind = "sym"` operator that `eigen`
can take directly. The function below builds it.

```{r jaccard-weighting}
jaccard_weighting <- function(data, ks, n_plots, min_cooccurrence) {
  P <- data$P[, ks, drop = FALSE]
  C <- Matrix::crossprod(P)                 # species x species co-occurrence counts; diag = occurrence
  f <- Matrix::diag(C)
  C <- methods::as(C, "TsparseMatrix")
  i <- C@i + 1L; j <- C@j + 1L; x <- C@x
  keep <- i != j & x > 0
  if (min_cooccurrence > 1L) keep <- keep & x >= min_cooccurrence
  i <- i[keep]; j <- j[keep]; x <- x[keep]
  jac <- x / (f[i] + f[j] - x)
  M <- Matrix::sparseMatrix(i = i, j = j, x = jac, dims = dim(C),
                            dimnames = dimnames(C))
  list(kind = "sym", M = M, species = data$species[ks])
}
```

`P` is the presence matrix restricted to the kept species. `C <-
Matrix::crossprod(P)` is the species by species co-occurrence count, where
`C[a, b]` counts the plots holding both `a` and `b` and the diagonal `C[a, a]` is
the occurrence count of `a`. The diagonal is pulled out as `f`, so `f[a]` is the
number of plots holding species `a`. Coercing `C` to a `TsparseMatrix` exposes
its stored entries as triplets: `C@i` and `C@j` are the zero-based row and column
indices, raised to one-based with `+ 1L`, and `C@x` is the count at each. The
`keep` mask drops the diagonal (`i != j`) and any zero count, and applies the
`min_cooccurrence` cutoff only on the surviving off-diagonal entries. The Jaccard
value `jac <- x / (f[i] + f[j] - x)` divides each co-occurrence count by the size
of the union of the two species' plot sets, since `f[a] + f[b] - C[a, b]` counts
plots holding either. Those values are reassembled into a sparse matrix with the
same dimensions and species names as `C`, and the function returns it with
`kind = "sym"`.

Two registrations wire it in. `register_weighting()` records the function under a
name, declaring its input shape and that it does not read cover.
`register_method()` pairs that weighting with the `eigen` factorization under a
friendly method name.

```{r jaccard-register}
register_weighting("jaccard", jaccard_weighting,
                   input = "species_species", supports_abundance = FALSE)
register_method("jaccard", weighting = "jaccard", factorization = "eigen",
                input = "species_species", native_output = "species")
"jaccard" %in% specvec_methods()
```

The method now lists alongside the built-ins and fits through the ordinary front
door. `species_embedding()` reads the method record, runs `jaccard_weighting` to
build the symmetric operator, checks that `eigen` accepts the `"sym"` kind it
produced, and takes the top eigenpairs.

```{r jaccard-fit}
emb <- species_embedding(x, method = "jaccard", dim = 6, min_occurrence = 2)
dim(emb$V)
nearest_species(emb, rownames(emb$V)[1], n = 3)
```

The embedding `$V` is a 12 by 6 matrix, one row per kept species, six dimensions
as requested. `eigen` accepts this operator because its `accepts` field is
`"sym"` and the weighting declared `kind = "sym"`: a symmetric matrix has a real
eigendecomposition, and the embedding is its leading eigenvectors scaled by the
square roots of the positive eigenvalues. `nearest_species()` reads the same
matrix and returns the species whose vectors point most nearly the same way,
which for this simulation are the species sharing a niche optimum.

The fitted object carries more than the matrix. `emb$method` records `"jaccard"`,
`emb$weighting` and `emb$factorization` record the pair that produced it, and
`emb$preprocessing` records the rare-species cutoffs and the species counts, so a
Jaccard embedding is self-describing in the same way an `abund_pmi` one is. The
new method gained every downstream consumer at once: anything that takes a
`specvec_embedding` (nearest neighbours, similarities, community pooling) works on
the Jaccard vectors without a line written for the new operand, because those
functions read the matrix and its metadata rather than the method that built it.

## The capability check

The kind contract is enforced at fit time. Pairing the Jaccard weighting,
which produces `"sym"`, with the `svd` factorization, which accepts `"implicit"`,
is a combination the engine refuses. Calling `species_embedding()` with
`method = NULL` and an explicit `weighting`/`factorization` pair forces exactly
that mismatch.

```{r capability-mismatch, error = TRUE}
species_embedding(x, weighting = "jaccard", factorization = "svd",
                  method = NULL, dim = 6, min_occurrence = 2)
```

The fitter runs the weighting, reads the `kind` of the operator it returned,
compares it against the factorization's `accepts`, and stops before any
decomposition when they differ. The message names both sides: `svd` consumes
`"implicit"`, but `jaccard` produced `"sym"`. The check is cheap and total. A
stored symmetric matrix has no matrix-vector closures for `svd` to call, and an
implicit operator has no `$M` for `eigen` to decompose, so a silent type error
later is converted into a clear refusal up front. The same gate is why every
built-in method pairs a weighting with a factorization that names the matching
kind.

Forcing the pair through the `weighting` and `factorization` arguments is the
explicit route a method record normally takes implicitly. When `method` is
supplied, the record names a weighting and a factorization that already agree on
kind, so the check passes by construction. Setting `method = NULL` and naming the
two halves directly is how an extension author tries a combination the registry
does not bless, and the check is what tells them, immediately, whether the
operand a weighting produces is the operand a factorization reads. A new
factorization that consumes `"sym"` would compose with `jaccard`, `pmi`, and
`abund_pmi` the moment it registered, with no change to any of them, because the
kind is the only thing the pairing depends on.

## Practical guidance

A few rules keep an extension clean. Add a weighting when the operand is new: a
different way to turn plots into a species by species (or plot by species) matrix,
as Jaccard turns co-occurrence into an overlap fraction. Add a factorization when
the operand is the same but the decomposition differs: a different way to read
vectors out of a matrix the existing weightings already produce. Most ecological
extensions are weightings, because the modelling choice usually lives in how
association is measured.

Match the kind to the factorization. A `"sym"` operator needs a symmetric matrix
and pairs with `eigen`; an asymmetric matrix breaks the real eigendecomposition
`eigen` assumes, so symmetrize it inside the weighting before returning. A
`"counts"` operator is the raw co-occurrence matrix that an objective
factorization such as `glove` consumes with its own loss. An `"implicit"`
operator carries matrix-vector closures for a factorization that never
densifies a large operand, the route correspondence analysis takes. Declaring a
kind the paired factorization does not accept fails at the capability check, by
design.

Keep the result deterministic. The `eigen` and `svd` factorizations return
vectors that are unique only up to sign, and the engine fixes each column's sign
so its largest-magnitude entry is positive. A weighting that returns a stable
operator therefore yields a stable embedding: two fits of the same data return
identical matrices. Randomness inside a weighting, an unordered set used as a key,
or a tie broken by floating-point noise would leak into the vectors and break that
guarantee, so a weighting that needs a random component should seed it. The sign
fix is applied after factorization, so an extension does not handle it; returning
a deterministic operator is enough.

Registrations belong in the `.onLoad` of an extending package, where they run
once when the package loads, or in a script sourced before fitting; the registry
is shared package state, not per-call configuration. Registering inside the
function that fits would re-register on every call, and a name registered twice
silently replaces the earlier entry, so a typo that shadows a built-in is easy to
make and quiet to miss. Keeping registrations in one place, run once, keeps the
set of available methods inspectable through `specvec_methods()` at any point.

Do not register a new method for what an argument on an existing weighting already
covers. A weighting that differs from a built-in only by a threshold or a switch
belongs as a parameter to that weighting. Two near-identical weightings split the
maintenance and tempt a caller to compare results that differ by one line of
hidden code; one weighting with an argument keeps the choice visible at the call.
The Jaccard weighting earned its own registration because the overlap fraction is
a genuinely different operand from PPMI, with its own bounded definition.

The built-in operators and the reasoning behind each are described in
`vignette("specvec-methods")`, which walks through the cover-weighted PMI default,
the compositional `clr` route, correspondence analysis, and the GloVe objective.
Reading what those operands measure is the fastest way to decide whether a new
modelling idea is a fresh weighting, an argument on an existing one, or a new
factorization over an operand SpecVec already builds.
