---
title: "Community embeddings and novelty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Community embeddings and novelty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(specvec)
```

A species embedding gives every species a vector. This vignette moves up one
level: it places each plot in that same space, compares plots to each other and
to a reference set, and scores how far a plot sits from familiar ground. The
worked examples run on a small simulated dataset where plots draw their species
from latent niches, so a plot's position in the embedding has a meaning we can
check against the way it was generated.

Three functions carry the work. `community_embedding()` pools species vectors
into one vector per plot. `community_similarity()` compares those plot vectors,
pairwise within one set or across two sets. `community_novelty()` scores each
plot by its distance to a reference body of communities. All three operate on the
same plot-by-dim matrix, so the path runs once from species vectors to plot
vectors and then branches into comparison and novelty without refitting.

The simulation builds `M` plots over a two-dimensional environment. Each species
has a niche centre, and a plot holds a species with probability that falls off
with the squared distance between the plot's environment and the species centre.
Cover is the relative niche match, scaled so the best-matched species in a plot
reads 100.

```{r sim}
sim_plots <- function(M = 300, S = 40, seed = 1) {
  set.seed(seed)
  mu <- matrix(rnorm(S * 2), S, 2)
  env <- matrix(rnorm(M * 2), M, 2)
  rows <- lapply(seq_len(M), function(p) {
    d2 <- rowSums((mu - matrix(env[p, ], S, 2, byrow = TRUE))^2)
    prob <- exp(-d2 / 2)
    present <- which(runif(S) < prob)
    if (length(present) < 2) present <- order(prob, decreasing = TRUE)[1:2]
    data.frame(plot = paste0("p", p),
               species = paste0("sp", present),
               cover = round(100 * prob[present] / max(prob[present]), 1))
  })
  do.call(rbind, rows)
}
```

## From species vectors to plot vectors

A species embedding `V` is a species-by-dim matrix: row `s` is the vector for
species `s`. A plot is a set of species, sometimes with cover. The community
embedding turns that set into a single point by pooling the vectors of the
species the plot contains, weighted by how much of each species is there:

```
U[p, ] = sum_s w[p, s] * V[s, ] / sum_s w[p, s]
```

The numerator adds up the species vectors, each scaled by its weight in the
plot; the denominator divides by the total weight, so `U[p, ]` is a weighted
average that lands inside the cloud of the plot's species. Two plots that share
most of their species, and weight them similarly, land near each other. A plot
made of niche specialists from one corner of the environment lands in that
corner of the embedding. The readout is the same regardless of which method
produced `V`, so plot vectors from different fits stay comparable as long as the
underlying species vectors agree.

The weighted average has a useful consequence. A plot vector is never an
arbitrary point; it is a convex combination of its species' vectors, so it sits
in the hull of the species it contains. A plot with one species sits exactly on
that species' vector. A plot with many species spread across the embedding sits
near their weighted centroid, which tends toward the middle of the cloud. The
geometry of the species embedding therefore propagates upward: if two species
sit close because they co-occur, plots built from either of them sit close too.
Species that the embedding never saw, because they fell under `min_occurrence`,
carry no vector and drop out of the pooling, so the plot vector is built from the
species that survived the rare-species filter.

The build starts with `specvec()`, which reads the long table into presence and
cover matrices. A species embedding follows, and `community_embedding()` pools
it into plot vectors.

```{r build}
df <- sim_plots(M = 300, S = 40, seed = 1)
x <- specvec(df, "plot", "species", abundance = "cover")
emb <- species_embedding(x, method = "abund_pmi", dim = 8, min_occurrence = 3)
comm <- community_embedding(x, embedding = emb)
comm
```

The print line reports the number of plots, the embedding dimension, the pooling
rule, and the provenance of the species vectors. The plot matrix `U` carries one
row per plot and one column per embedding dimension.

```{r dims}
dim(comm$U)
head(rownames(comm$U), 4)
```

With 300 plots and `dim = 8`, `U` is 300 by 8. Each row is a plot's coordinates
in the eight-dimensional species space, ready for the similarity and novelty
tools below.

## Cover vs presence pooling

The weight `w[p, s]` is where cover enters. With `weights = "cover"` the weight
is the plot's cover for species `s`, so a species covering 80 of a plot pulls the
plot vector four times as hard as one covering 20. With `weights = "presence"`
every present species carries weight 1, and the plot vector is the plain mean of
the vectors of the species on the list. The two share the same species
embedding; they differ only in how they average it. The same `emb` flows into
both calls below, so any difference in the resulting plot positions comes from
the weighting rule alone, which makes the comparison a clean A/B test of cover
against presence.

```{r pooling}
comm_cov <- community_embedding(x, embedding = emb, weights = "cover")
comm_pre <- community_embedding(x, embedding = emb, weights = "presence")
comm_cov$pooling
comm_pre$pooling
```

The two plot matrices are built from the same `V` and the same species lists,
yet they place the plots differently. A quick way to see this is the per-plot
distance between the cover vector and the presence vector: zero would mean the
two agree, and most plots sit well above zero.

```{r pooling-diff}
delta <- sqrt(rowSums((comm_cov$U - comm_pre$U)^2))
summary(delta)
```

The gap is largest in plots with one dominant species and several trace
companions. Cover pooling pulls such a plot toward the dominant's vector;
presence pooling treats the dominant and the traces alike and sits closer to the
centroid of the whole species list. In a plot where one species covers 90 and
two others cover 5 each, cover pooling weights the dominant eighteen times as
heavily as either companion, so the plot vector lands almost on the dominant's
position. Presence pooling weights all three equally and lands a third of the way
between them. The two readings of the same plot answer different questions about
what the plot is.

The choice follows the question. Cover pooling suits abundance-driven analyses,
where a plot carpeted by one species should read as that species' kind of place,
and where the cover values come from a consistent scale. Presence pooling suits
species-list work, where the data are checklists with no reliable abundance, or
where every recorded species should count once regardless of how much of it was
there. When the input has no cover column, `community_embedding()` falls back to
presence pooling on its own and reports that it did so, so a presence-only
dataset behaves sensibly without any extra argument.

The scatter below shows the plots in the first two community dimensions under
cover pooling, with four plots labelled. Position carries the latent niche: plots
near each other share their abundant species.

```{r pooling-plot, fig.width = 7, fig.height = 5}
set.seed(2)
lab <- sample(rownames(comm_cov$U), 4)
plot(comm_cov$U[, 1], comm_cov$U[, 2],
     pch = 16, col = "grey60", cex = 0.7,
     xlab = "community dim 1", ylab = "community dim 2",
     main = "Plots in community space (cover pooling)")
points(comm_cov$U[lab, 1], comm_cov$U[lab, 2], pch = 16, col = "black")
text(comm_cov$U[lab, 1], comm_cov$U[lab, 2], labels = lab, pos = 3, cex = 0.8)
```

## Fitting fresh vs reusing an embedding

`community_embedding()` can fit a species embedding on the spot. Passing
`method`, `dim`, and the rare-species filters without an `embedding` argument
runs `species_embedding()` internally and pools the result.

```{r fit-fresh}
comm_fresh <- community_embedding(x, method = "abund_pmi", dim = 8,
                                  min_occurrence = 3)
comm_fresh
```

The alternative is to fit once and pass the fitted object through `embedding =`.
When the internal fit uses the same method, dimension, and filters as a pre-fit
embedding, the two paths produce the same species vectors and therefore the same
plot matrix `U`.

```{r fit-reuse}
comm_reuse <- community_embedding(x, embedding = emb)
max(abs(comm_fresh$U - comm_reuse$U))
```

The maximum element-wise difference is zero to numerical precision: the fresh fit
and the reused embedding describe the same plots in the same coordinates. The
equality holds because the species embedding is deterministic. SpecVec fixes the
sign of every embedding column, so two fits of the same data with the same
filters return byte-identical species vectors, and pooling them gives byte
-identical plot vectors. Reusing one embedding has a practical payoff. Several
community sets pooled from one frame share an axis system, so their plot vectors
compare directly and the fit runs once instead of once per set. The reuse path
also lets the species embedding come from a larger or more authoritative dataset
than the plots being pooled, which is the usual setup when a small survey is
placed in the coordinate system of a continental reference.

`normalize = TRUE` rescales each plot vector to unit length after pooling, which
sends every plot to the surface of a sphere and makes cosine comparisons depend
on direction alone. Leave it off when distances between plots should reflect both
direction and length; turn it on when only the compositional direction matters.

```{r normalize}
comm_norm <- community_embedding(x, embedding = emb, normalize = TRUE)
comm_norm$normalized
range(sqrt(rowSums(comm_norm$U^2)))
```

## Community similarity

`community_similarity()` compares plot vectors. Called on one community object it
returns the self-similarity matrix: every plot against every other plot. With
`metric = "cosine"` the entries are cosine similarities in `[-1, 1]`, where 1 is
two plots pointing the same way in species space and a value near zero is two
plots with little in common.

```{r sim-self}
S <- community_similarity(comm_cov, metric = "cosine")
dim(S)
round(S[1:4, 1:4], 2)
```

The diagonal is 1, each plot is identical to itself. An off-diagonal cell close
to 1 marks a pair of plots whose abundant species overlap and point the same
direction; a cell near zero or below marks a pair drawing on different niches. In
the submatrix above, the larger cells flag plot pairs that share their dominant
species. Cosine reads composition through the embedding rather than through raw
species overlap, so two plots can score high even with no species in common, as
long as their species occupy nearby parts of the embedding. That is the
difference from a Jaccard or Bray-Curtis index computed on the species lists:
the embedding lets ecologically similar species stand in for each other, while a
set-overlap index counts only the species that match by name.

A reference set goes in through the `reference` argument. The result has one row
per plot in `object` and one column per plot in `reference`, so it answers "how
does each of these plots compare with each of those". Splitting the plots into
two halves gives a worked cross-comparison.

```{r sim-ref}
plots_all <- rownames(comm_cov$U)
half <- length(plots_all) %/% 2
xa <- specvec(df[df$plot %in% plots_all[1:half], ], "plot", "species",
              abundance = "cover")
xb <- specvec(df[df$plot %in% plots_all[(half + 1):length(plots_all)], ],
              "plot", "species", abundance = "cover")
ca <- community_embedding(xa, embedding = emb)
cb <- community_embedding(xb, embedding = emb)
Scross <- community_similarity(ca, reference = cb, metric = "cosine")
dim(Scross)
```

`metric = "euclidean"` switches the readout to straight-line distance between
plot vectors. Distance grows as plots diverge, so it reads as a dissimilarity:
small values are near neighbours, large values are far apart. Cosine ignores
vector length and cares about direction; Euclidean distance feels both, and the
two rank plot pairs differently when plots vary in how strongly their species
load on the embedding. A species-poor plot and a species-rich plot built from
the same niche point the same way and score high on cosine, yet sit a real
distance apart in Euclidean terms because their vectors have different lengths.
Pairing `metric = "euclidean"` with `normalize = TRUE` at the pooling step
removes the length difference, so the distance then tracks direction alone and
the two metrics agree on ordering. The Euclidean distance from
`community_similarity()` is the same distance `community_novelty()` averages over
its nearest neighbours, so the two functions read plots through one geometry.

```{r sim-euclid}
D <- community_similarity(comm_cov, metric = "euclidean")
round(D[1:4, 1:4], 2)
```

An `image()` of a small cosine block shows the structure at a glance. Brighter
cells are more similar plot pairs; the bright diagonal is each plot against
itself.

```{r sim-image, fig.width = 5.5, fig.height = 5}
sub <- S[1:20, 1:20]
image(seq_len(20), seq_len(20), sub,
      col = hcl.colors(12, "YlOrRd", rev = TRUE),
      xlab = "plot", ylab = "plot",
      main = "Cosine similarity (first 20 plots)")
```

## Community novelty

Novelty asks a different question. It measures how far a plot sits from a body of
known communities. `community_novelty()` scores each plot in `object` by its mean
Euclidean distance to the `k` nearest plots in `reference`.
A plot whose `k` nearest reference neighbours are all close scores low; a plot
that sits far from everything in the reference set scores high. The score is a
distance, so it has no upper bound and is read relative to the spread of the
reference.

The mean-of-nearest-neighbours definition matters. A plot's novelty is the
average distance to its `k` closest reference plots. That choice sits between two
cruder options: distance to the single closest reference plot, which reacts to
one chance neighbour, and distance to the reference centroid, which ignores local
structure. Averaging over a handful of neighbours keeps the score steady against
a lone reference plot that happens to sit nearby, so a query plot with one close
neighbour and four distant ones still scores as novel. The score reads as
distance to a local pocket of the reference, which is the property that lets it
separate plots in dense regions from plots in sparse ones.

The setup splits the plots into a reference set and a query set, both pooled from
the same embedding so their coordinates align. The reference is the bulk of the
simulated plots; the query mixes ordinary plots with a handful built to land in
an unusual part of the species space.

```{r novelty-setup}
ref_plots <- plots_all[1:250]
qry_plots <- plots_all[251:300]
x_ref <- specvec(df[df$plot %in% ref_plots, ], "plot", "species",
                 abundance = "cover")
comm_ref <- community_embedding(x_ref, embedding = emb)
```

To force genuine novelty, build a few odd plots that pair the species sitting
farthest apart in the embedding. Such species rarely co-occur in the reference,
so a plot holding both pools to a point between the reference clusters, in a
thinly populated region.

```{r novelty-odd}
dmat <- as.matrix(dist(emb$V))
ut <- which(upper.tri(dmat), arr.ind = TRUE)
ord <- order(dmat[ut], decreasing = TRUE)
set.seed(7)
odd <- do.call(rbind, lapply(1:6, function(i) {
  pr <- ut[ord[i], ]
  data.frame(plot = paste0("odd", i), species = emb$species[pr],
             cover = round(runif(2, 40, 100), 1))
}))
qry_df <- rbind(df[df$plot %in% qry_plots, ], odd)
x_qry <- specvec(qry_df, "plot", "species", abundance = "cover")
comm_qry <- community_embedding(x_qry, embedding = emb)
```

With reference and query in hand, `community_novelty()` returns one score per
query plot. `k = 5` averages over the five nearest reference communities, which
smooths over a single unusually close or far neighbour while still tracking the
local density of the reference.

```{r novelty-run}
nov <- community_novelty(comm_qry, comm_ref, k = 5)
round(sort(nov, decreasing = TRUE)[1:8], 3)
```

The six odd plots take the top six places. They pair species the reference rarely
holds together, so their plot vectors land in a thinly populated region and their
nearest five reference neighbours sit far away, around 0.5 to 0.7 in embedding
distance. Ordinary query plots, built the same way as the reference, score below
0.3 because they fall inside the reference cloud where neighbours are close.

```{r novelty-check}
is_odd <- grepl("^odd", names(nov))
c(odd = mean(nov[is_odd]), ordinary = mean(nov[!is_odd]))
```

The mean novelty of the deliberately odd plots runs about four to five times the
ordinary query plots, which is the behaviour the score is built to give. A
histogram shows the bulk of plots at low novelty with a tail of outliers; the odd
plots sit in that tail, marked by the red rug.

```{r novelty-hist, fig.width = 7, fig.height = 4}
hist(nov, breaks = 12, col = "grey80", border = "white",
     xlab = "novelty (mean distance to 5 nearest reference plots)",
     main = "Distribution of community novelty")
rug(nov[is_odd], col = "red", lwd = 2)
```

The choice of `k` trades sensitivity against stability. Small `k` (1 or 2) reacts
to the single closest reference plot and flags anything off the beaten track, at
the cost of noise. Larger `k` averages over a neighbourhood and reports distance
to a region rather than a point, which is steadier when the reference is dense.
The function caps `k` at the number of reference communities, so passing a `k`
larger than the reference simply uses every reference plot. The reference may be
a `specvec_community` or a plain numeric matrix of community vectors, so a fixed
set of reference coordinates, saved once, can be reused across many query
batches without rebuilding a community object each time.

One detail keeps the comparison honest: the query and the reference must live in
the same coordinate system. Pooling both through the same species embedding `emb`
guarantees that. A novelty score computed against a reference pooled from a
different fit would mix two axis systems and report distances that mean nothing,
which is the same reason the similarity examples above pooled their two halves
from one shared `emb`.

## Practical guidance

A few rules of thumb cover most uses.

Choose pooling by what the data mean. Use `weights = "cover"` when abundance is
reliable and a dominant species should dominate the plot vector, so a stand
carpeted by one species reads as that species' kind of place. Use
`weights = "presence"` for checklists, for occurrence-only surveys, and whenever
every recorded species should count once. The default falls back to presence on
its own when the data carry no cover column.

Normalize when only direction matters. Leaving `normalize = FALSE` keeps both the
direction and the length of each plot vector, so a strongly loaded plot reads as
more extreme than a faint one. Set `normalize = TRUE` when plots should be
compared on composition alone, which is the common choice ahead of cosine
clustering, since every plot then sits on the unit sphere and length drops out.

Pool several community sets from one shared embedding. Fitting the species
embedding once and passing it through `embedding =` to each
`community_embedding()` call guarantees the sets share an axis system. Plot
vectors from different sets then compare directly, and the fit cost is paid once.
Refitting per set risks sign or scaling drift that breaks cross-set comparison.

Scale `k` to the reference. With a small reference set a large `k` averages over a
big fraction of all reference plots and washes out the local structure novelty is
meant to read; `community_novelty()` already caps `k` at the number of reference
communities. With a few hundred reference plots, `k` of 5 to 10 balances
sensitivity against noise. Keep `k` well below the reference size so the score
still reflects a local neighbourhood.

Give novelty a meaningful reference. The score is a distance to whatever
reference you pass, so a reference that is itself narrow makes ordinary plots look
novel, and a reference that already contains the odd communities makes them look
ordinary. A reference assembled to represent "what is known" gives a score that
reads as "how far from the known". This is also the limit to keep in mind:
novelty is relative to the reference, not an absolute outlier score, so a plot
flagged as novel against one reference can be unremarkable against another.
