---
title: "Alien integration trajectories"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alien integration trajectories}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

A neophyte arrives in disturbed ground with a ruderal flora and, over decades, may
settle into the resident native community. `integration_trajectory()` measures
that settling: it tracks the focal species' distance to the native-community
centroid through time, in one fixed embedding frame. A falling distance is
integration -- the neophyte's associates shift from where it arrived toward the
native community.

```{r setup}
library(specvec)
```

## The construction

Everything is read out in a single fixed frame, so the points compare across time
without alignment (the fixed-frame idea from `vignette("specvec-methods")` and
`species_trajectory()`):

- The **frame** is fitted once on the resident species pool. The focal neophyte
  is placed per time window at the cover-weighted centroid of the species it
  co-occurs with, exactly as `species_trajectory()` does.
- The **native community** is placed per window at the centroid of its pooled
  plot embeddings, in the same frame.
- The trajectory is the distance between the two per window.

## A simulated invasion

The simulation holds a native community as a stable backbone in every decade,
and lets a focal `"alien"` ride a ruderal flora early and the native flora late,
so its co-occurrents drift from one to the other.

```{r sim}
sim_invasion <- function(seed = 1) {
  set.seed(seed)
  native  <- paste0("nat", 1:8)
  ruderal <- paste0("rud", 1:8)
  decades <- c(1990, 2000, 2010, 2020)
  rows <- list(); pid <- 0L
  for (d in seq_along(decades)) {
    share <- (d - 1) / (length(decades) - 1)      # the neophyte's pull toward natives
    for (i in 1:120) {                            # plots carrying the neophyte
      pid <- pid + 1L
      pool <- if (runif(1) < share) native else ruderal
      sp <- c(sample(pool, sample(3:6, 1)), "alien")
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decades[d])
    }
    for (i in 1:60) {                             # resident native backbone
      pid <- pid + 1L
      sp <- sample(native, sample(3:6, 1))
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decades[d])
    }
  }
  do.call(rbind, rows)
}

df <- sim_invasion()
x  <- specvec(df, "plot", "species", abundance = "cover", time = "decade")
x
```

Trace the neophyte against the native species:

```{r run}
it <- integration_trajectory(
  x, species = "alien",
  native = paste0("nat", 1:8),     # the community it integrates into
  dim = 16, min_occurrence = 3
)
it
```

The distance falls decade by decade as the neophyte's co-occurrents shift into the
native community:

```{r plot}
d <- as.data.frame(it)
plot(d$center, d$distance, type = "b", pch = 19,
     xlab = "decade", ylab = "distance to native centroid",
     main = "alien integration over time")
```

## Frame and native are two different sets

`integration_trajectory()` separates the coordinate system from the target.

- `frame` defaults to every species except the focal one -- the whole resident
  pool. Keeping the ruderal pole in the frame is what makes an early position well
  defined: the neophyte starts among the flora it arrived with, and the motion
  toward the natives is a real displacement in the space.
- `native` is the subset whose centroid is the target. It defaults to the whole
  frame, but for a real measurement you name the native species, so the distance
  is measured against the community you care about.

Fitting the frame on the native subset alone would erase the ruderal pole, and the
early position would have nothing to sit at.

## Anchor on resampled plots

Across decades, raw plot data confounds integration with where and what was
sampled each decade: a shift in the distance could be a shift in the survey, not
in the species. Restrict the data to **resampled (ReSurvey) plots** before
building the `specvec` object, so the same locations are followed through time.
`integration_trajectory()` takes whatever plots it is given and does not hard-code
a survey-design column; the restriction is yours to apply.

## On your own data

A European Vegetation Archive (EVA) export carries the columns this needs: a
plot id, a taxon, a cover, a decade, a native/neophyte status, and a ReSurvey
flag. The native set comes from the status column; the ReSurvey flag selects the
resampled plots.

```{r eva, eval = FALSE}
library(data.table)
sp <- fread("species_export.csv")
hd <- fread("header_export.csv")                 # plot-level: decade, ReSurvey flag

resurvey <- hd[`ReSurvey plot (Y/N)` == "Y", PlotObservationID]
sp <- sp[PlotObservationID %in% resurvey]        # follow the same locations

x <- specvec(sp, plot = "PlotObservationID", species = "taxon",
             abundance = "cover", time = "Decade")

natives <- unique(sp[STATUS == "native", taxon])
integration_trajectory(x, species = "Robinia pseudoacacia",
                       native = natives, min_occurrence = 5)
```

The package carries no vegetation data of its own; the trajectory is meant to run
on yours.
