---
title: "Tracking species and communities through time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracking species and communities through time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(specvec)
```

Vegetation data often arrives stamped with time: plots surveyed across decades,
a ReSurvey campaign returning to the same locations, a chronosequence read off
land-use history. The natural question is whether a species or a community has
moved through compositional space, and where it went. SpecVec answers this with
two front-door verbs, `species_trajectory()` and `community_trajectory()`, both
built on one idea: fit a single coordinate system once, then read every time
window out in that system.

## The cross-time alignment problem

The obvious approach is to embed each decade on its own. Fit one species
embedding on the 1990s plots, another on the 2010s plots, and compare the two
coordinate matrices. The approach fails for a numerical reason. An eigendecomposition
or singular value decomposition returns vectors
that are unique only up to sign, and when eigenvalues are close, up to rotation
within the degenerate subspace. SpecVec fixes each column's sign deterministically
(see `vignette("specvec-methods")`), so refitting the *same* data twice gives the
same matrix. Two *different* fits carry no such guarantee. The axes of the 1990s
embedding and the axes of the 2010s embedding are separate bases for separate
matrices. A species sitting at coordinate `(1.2, -0.4)` in one fit and
`(1.2, -0.4)` in the other has not necessarily stayed put; the two coordinate
pairs are written in different languages.

People patch this with Procrustes rotation: find the orthogonal transform that
best maps one fit onto the other, apply it, then compare. That works when the
two communities are mostly the same and only a few species moved. It breaks
exactly when the data are interesting. Procrustes assumes a shared structure to
align to, so if the species pool itself turned over between decades, the
alignment absorbs the very signal being measured, and a real compositional shift
gets rotated away. The rotation is also estimated from the data it then judges,
which is circular.

The fixed-frame solution sidesteps all of this. One *frame* embedding is fitted
once, on a stable reference set of species. Each time window is then projected
into that frame without refitting: a focal species is placed, per window, at the
cover-weighted centroid of the frame species it co-occurs with in that window.
The frame never moves, so there is nothing to align. A point in 1990 and a point
in 2010 are coordinates in the *same* basis, and their difference is a genuine
displacement. There is no per-window rotation, no Procrustes step, and no
circularity, because the thing measured (the focal species' position) is never
part of the thing that defines the axes (the frame).

The data that drives this needs a time column, supplied to `specvec()` at build
time. The simulation below codes four decades. A focal species rides one species
pool early and drifts toward a second pool late, while a backbone of unrelated
plots holds steady throughout.

```{r sim}
sim_time <- function(seed = 1) {
  set.seed(seed)
  a <- paste0("a", 1:8); b <- paste0("b", 1:8)
  decs <- c(1990, 2000, 2010, 2020)
  rows <- list(); pid <- 0L
  for (di in seq_along(decs)) {
    share <- (di - 1) / (length(decs) - 1)
    for (i in 1:100) {
      pid <- pid + 1L
      pool <- if (runif(1) < share) b else a
      sp <- c(sample(pool, sample(3:6, 1)), "focal")
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decs[di])
    }
    for (i in 1:50) {
      pid <- pid + 1L
      sp <- sample(a, sample(3:6, 1))
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decs[di])
    }
  }
  do.call(rbind, rows)
}

x <- specvec(sim_time(), "plot", "species", abundance = "cover", time = "decade")
x
```

The `focal` species starts among the `a` pool and is pulled toward the `b` pool
as the decades pass. The `a`-only backbone plots keep the `a` species anchored,
so they make a sensible frame.

## A single window

A frame is itself just a species embedding. To see what one window looks like in
isolation, fit an embedding on a slice of the data with the `time` argument.
A numeric vector selects a closed interval, so `time = c(1990, 2000)` trains on
every plot whose decade falls between 1990 and 2000.

```{r window}
emb_early <- species_embedding(x, time = c(1990, 2000), dim = 6, min_occurrence = 2)
emb_early
```

This is a valid embedding of the early decades. The trouble is the second one.

```{r window2}
emb_late <- species_embedding(x, time = c(2010, 2020), dim = 6, min_occurrence = 2)
emb_late
```

`emb_early` and `emb_late` are two independent fits. Each has its own sign-fixed
axes, but those axes were chosen against different co-occurrence operators, so the
6 dimensions of one do not correspond to the 6 dimensions of the other. Comparing
`emb_early$V` to `emb_late$V` entry by entry measures the difference between two
coordinate systems as much as any difference between two decades. This is the
alignment problem in miniature, and it is the reason `species_trajectory()`
exists: it fits the frame once and reuses it for every window, so the comparison
is clean by construction.

## Quantifying the alignment problem

The argument above is easy to state and easy to underweight. It helps to put a
number on it. We take the species shared between the two independent window fits,
read off the displacement each species appears to make from the early fit to the
late fit, and then compare that naive displacement against the displacement the
fixed frame reports for the same species. The two readings should disagree
sharply, because the naive one mixes a genuine compositional shift with a change
of basis that nobody asked for.

We start with the shared species. Both fits keep every species above the
occurrence floor in their own window, so the overlap is the set of species
present in both halves of the series.

```{r align-shared}
Ve <- emb_early$V; Vl <- emb_late$V
shared <- intersect(rownames(Ve), rownames(Vl))
shared <- setdiff(shared, "focal")
length(shared)
```

The naive displacement is the row-wise Euclidean distance between the early and
late coordinate matrices, restricted to those shared species. If the two fits
spoke the same language, a species that did not move would sit at distance zero
and one that moved a little would sit at a small distance. They do not speak the
same language, so the distances are large and roughly uniform across species that
the simulation never perturbed.

```{r align-naive}
naive <- sqrt(rowSums((Ve[shared, ] - Vl[shared, ])^2))
summary(naive)
```

These numbers are not a measurement of movement. They are the gap between two
arbitrary bases, the same gap that a Procrustes step would try to rotate away.
The honest test is to grant the comparison its best chance and fit the optimal
orthogonal rotation that maps the late fit onto the early fit, then measure what
displacement survives. The rotation is the closed-form orthogonal Procrustes
solution from the singular value decomposition of the cross-product.

```{r align-flip}
E <- Ve[shared, ]; L <- Vl[shared, ]
sv <- svd(crossprod(L, E))
R <- sv$u %*% t(sv$v)
L_rot <- L %*% R
aligned <- sqrt(rowSums((E - L_rot)^2))
summary(aligned)
```

Even after the best possible rotation a sizeable residual survives, and the
rotation was fitted to the very species it then scores, which is the circularity
the alignment cannot escape. There is no clean displacement to extract here.

The fixed frame reports something different. We fit one frame on all species,
then read each shared species at its early and late window centroid in that one
frame (a focal species never enters its own centroid), and take the same row-wise
distance. Because the frame never moved, this distance is a displacement and
nothing else.

```{r align-frame}
frame_all <- species_embedding(x, dim = 6, min_occurrence = 2)
tr_shared <- species_trajectory(x, species = shared, frame_embedding = frame_all,
                               by = c(1985, 2005, 2025))
ds <- as.data.frame(tr_shared)
e <- ds[ds$center == 1995, ]; l <- ds[ds$center == 2015, ]
l <- l[match(e$species, l$species), ]
dd <- paste0("d", 1:6)
frame_disp <- sqrt(rowSums((as.matrix(e[, dd]) - as.matrix(l[, dd]))^2))
summary(frame_disp)
```

The fixed-frame displacements are smaller and, more to the point, interpretable.
The species the simulation held steady move little, and the spread reflects real
differences in how their associates turned over. The naive median sits well above
the fixed-frame median, and that gap is the alignment problem made numeric.

```{r align-compare}
c(naive_median = median(naive),
  frame_median = median(frame_disp),
  ratio = median(naive) / median(frame_disp))
```

The ratio is the size of the artifact the fixed frame removes. The naive
comparison spends most of its magnitude on the change of basis; the fixed-frame
comparison spends all of it on the data. This is what we mean when we call the
per-window-refit approach uninterpretable: the arithmetic is correct and the
coordinates it runs on are arbitrary.

A reader might ask whether a better alignment, a full Procrustes rotation rather
than the greedy axis match used above, would close the gap. It would shrink the
residual, and on data where the two windows share almost all their species and
structure it can shrink it a long way. The trouble is that the rotation is fitted
to make the two coordinate matrices agree, so it absorbs whatever it can,
including genuine movement, and the better it fits the more real signal it can
quietly remove. On a series where the species pool turns over, which is the case
worth studying, the alignment has the most license to overcorrect exactly where
the most is happening. The fixed frame avoids the question by never asking it.
Nothing is rotated, so nothing real can be rotated away, and the cost is only that
the frame must be fitted on a set of species stable enough to anchor every
window. The frame is a modelling decision made once and in the open. The
per-window rotation is re-estimated silently at every comparison.

## Species trajectories

`species_trajectory()` traces one or more focal species through the fixed frame.
The frame defaults to every species except the focal ones, so the focal species
move through a background they do not help define. Each window places the focal
species at the cover-weighted centroid of the frame species it co-occurs with in
that window, and a focal species never contributes to its own position.

The exclusion is what makes the trajectory clean. If the focal species helped
define the frame, its own position would be partly a function of itself, and a
track that moved would be hard to separate from a track that re-weighted its own
contribution to the axes. Holding the focal species out of the frame makes the
centroid a direct readout: the focal species is wherever its current associates
already sit. When those associates turn over, the focal species moves; when they
hold, it stays. Nothing about the axes changes between windows, so the only thing
that can change a coordinate is the composition the focal species keeps. That is
the property that makes a difference between two window positions a measurement
the data supports.

```{r traj}
tr <- species_trajectory(x, species = "focal", by = NULL, dim = 6,
                         min_occurrence = 2)
tr
```

The print shows the frame (here the 16 non-focal species, embedded once), the
four windows, and the per-cell support range. With `by = NULL`, one window is cut
per distinct decade. The tidy form spreads the array into one row per
focal-species-by-window cell, carrying the window center, the support, and the
`dim` coordinate columns `d1` through `d6`.

```{r traj-df}
d <- as.data.frame(tr)
d
```

The `support` column is the number of window plots that actually contain the
focal species, and `center` is the numeric window center used for ordering and
plotting. The `windows` table records the same window metadata directly, with the
plot count per window.

```{r traj-windows}
tr$windows
```

A cell is `NA` in the coordinate columns when the focal species co-occurs with no
frame species in that window, so its centroid is undefined. That happens in
sparse windows, and `as.data.frame(tr, na.rm = TRUE)` drops those rows. Here the
focal species occurs in roughly 100 plots per decade, so every cell is backed by
ample co-occurrence and none are `NA`.

The displacement of the focal species between consecutive windows is the
Euclidean distance between successive rows of the coordinate matrix. The frame is
fixed, so this distance is a real measure of how far the species' associates
moved, not an artifact of refitting. A displacement near zero means the focal
species kept the same company across the two windows; a large displacement means
its co-occurrents shifted toward a different region of the frame. Because the
distance is taken across all `dim` coordinates at once, it captures movement
along every axis together, which matters when a turnover spreads across several
frame dimensions at the same time.

```{r traj-plot}
dcols <- grep("^d[0-9]+$", names(d))
co <- as.matrix(d[order(d$center), dcols])
step <- sqrt(rowSums((co[-1, ] - co[-nrow(co), , drop = FALSE])^2))
mids <- (d$center[order(d$center)][-1] + d$center[order(d$center)][-nrow(co)]) / 2

plot(mids, step, type = "b", pch = 19,
     xlab = "decade midpoint", ylab = "displacement in frame",
     main = "focal species: movement between successive decades")
```

The displacement is largest across the middle decades, where the simulated pull
from the `a` pool to the `b` pool is steepest, and smaller at the ends. The first
coordinate alone tells a similar story: tracking `d1` against the window center
shows the focal species sliding along the frame's leading axis as its co-occurrents
turn over.

```{r traj-d1}
plot(d$center, d$d1, type = "b", pch = 19,
     xlab = "decade", ylab = "first frame coordinate (d1)",
     main = "focal species: position on the leading frame axis")
```

Both views read the same fixed frame, so a movement on the plot is a movement in
the data. The displacement plot and the `d1` trace answer slightly different
questions. The displacement plot asks how fast the species is moving between
consecutive windows, with no sign and no direction, so it peaks where change is
quickest. The `d1` trace asks where the species sits on the dominant axis at each
window, with a direction, so it shows the cumulative drift. A monotone `d1` trace
with a single displacement peak in the middle is the signature of a steady
directional shift that runs fastest halfway through, which is exactly the pull the
simulation built in.

## Tracing several species at once

`species` takes a vector, so one call traces several focal species through the
same frame. This is more than a convenience. When every focal species reads the
same fixed frame, their trajectories are directly comparable to each other as
well as across time, which a series of separate single-species calls could not
guarantee if each refitted its own background. We add two more focal species to
the simulation to have something to trace alongside the original.

```{r multi-sim}
df0 <- sim_time()
extra <- do.call(rbind, lapply(split(df0, df0$plot), function(g) {
  if (runif(1) < 0.4) rbind(g, transform(g[1, ], species = "rider", cover = 30))
  else if (runif(1) < 0.4) rbind(g, transform(g[1, ], species = "stayer", cover = 30))
  else g
}))
xm <- specvec(extra, "plot", "species", abundance = "cover", time = "decade")
```

With three focal names the return array gains a leading dimension. `U` is now a
focal-by-window-by-dim array, three slabs deep, one slab per focal species, each
slab a window-by-dim matrix in the shared frame. The `support` matrix gains a
matching row per focal species. The array layout keeps the three indices
separate so that any one of them can be held fixed and the others read out: fix a
species to get its full window-by-dim track, fix a window to compare where all
three species sit at one decade, or fix a dimension to watch every species move
along a single frame axis. The tidy data frame flattens the first two indices
into rows and keeps the dimensions as columns, which is the shape most plotting
and modelling code expects, while the raw array stays available for the slicing a
data frame makes awkward.

```{r multi-traj}
trm <- species_trajectory(xm, species = c("focal", "rider", "stayer"),
                         by = NULL, dim = 6, min_occurrence = 2)
dim(trm$U)
```

The tidy form spreads that array into one row per focal-species-by-window cell,
so three species across four windows give twelve rows, carrying the `species`
column to tell them apart.

```{r multi-df}
dm <- as.data.frame(trm)
table(dm$species)
head(dm[, c("species", "window", "support", "d1")], 6)
```

Plotting the first frame coordinate for two of the focal species on one pair of
axes shows their tracks side by side in the same frame. The species the
simulation pulls toward the `b` pool should sweep across the leading axis while a
species spread evenly across plots should sit nearly flat.

```{r multi-plot}
f <- dm[dm$species == "focal", ]; s <- dm[dm$species == "stayer", ]
rng <- range(c(f$d1, s$d1))
plot(f$center, f$d1, type = "b", pch = 19, ylim = rng,
     xlab = "decade", ylab = "first frame coordinate (d1)")
lines(s$center, s$d1, type = "b", pch = 17, lty = 2)
legend("topright", c("focal", "stayer"), pch = c(19, 17), lty = c(1, 2), bty = "n")
```

The two traces read off one frame, so the vertical gap between them at any decade
is a real difference in where the two species sit relative to the same
background.

## Choosing windows

The `by` argument controls how the time axis is split. The default `by = NULL`
makes one window per distinct time value, which suits decade-coded data where the
decades are the natural units.

```{r win-default}
tr$windows
```

Four decades, four windows, each centered on its decade and holding 150 plots
(100 carrying the focal species plus 50 backbone). Passing a numeric break vector
instead bins the time axis with `cut()`, using closed lower bounds. The breaks
`c(1985, 2005, 2025)` cut two bins, an early bin spanning 1985 to 2005 and a late
bin spanning 2005 to 2025.

```{r win-breaks}
tr2 <- species_trajectory(x, species = "focal", by = c(1985, 2005, 2025),
                          dim = 6, min_occurrence = 2)
tr2$windows
```

The window centers are now the midpoints of the break intervals, 1995 and 2015,
and each window pools two decades, so `n_plots` is 300 rather than 150. Coarser
windows hold more plots per window, which steadies the centroid in each cell while
lowering temporal resolution. Both readings describe the same trajectory at two
grains, and the break vector sets the grain.

## Reusing one frame across analyses

Both verbs fit a frame embedding internally, and both accept a pre-fitted one
through `frame_embedding =`. Passing the same frame to a species trajectory and a
community trajectory puts the two readings in one coordinate system, so a
species' track and the community novelty around it are written in the same basis
and can be overlaid without any further alignment. We fit the frame once with
`species_embedding()`, on a stable backbone, and hand it to both verbs.

```{r reuse-fit}
frame <- species_embedding(x, dim = 6, min_occurrence = 2)
frame
```

The species trajectory now reuses this frame instead of fitting its own. The
focal species rides through the supplied coordinate system, and the result is
byte-comparable to anything else that reads the same frame.

```{r reuse-species}
tr_f <- species_trajectory(x, species = "focal", frame_embedding = frame,
                          by = NULL)
identical(tr_f$frame$V, frame$V)
```

The community trajectory takes the same frame, so its per-window communities and
its novelty live in the basis the species track already used.

```{r reuse-comm}
ct_f <- community_trajectory(x, frame_embedding = frame, by = NULL, k = 5)
identical(ct_f$frame$V, frame$V)
```

The shared frame is what lets `integration_trajectory()` read a neophyte's
distance to a native community over time as a single comparable number. The
neophyte's position and the community's position both come out of one fixed
frame, so their separation is a distance in that frame and not a mismatch between
two bases.

There is a subtler use. A species can be projected into a frame that does not
contain it, because the projection only reads the frame species the focal
co-occurs with. We fit a frame on every species except the focal, drop the focal
from it explicitly so the frame genuinely never saw it, and place the focal into
that frame anyway.

```{r reuse-outframe}
others <- setdiff(x$species, "focal")
frame_no_focal <- species_embedding(x, dim = 6, min_occurrence = 2)
keep <- rownames(frame_no_focal$V) %in% others
frame_no_focal$V <- frame_no_focal$V[keep, , drop = FALSE]
frame_no_focal$species <- rownames(frame_no_focal$V)
"focal" %in% frame_no_focal$species
tr_out <- species_trajectory(x, species = "focal",
                            frame_embedding = frame_no_focal, by = NULL)
sum(is.na(tr_out$U))
```

The frame contains no `focal` row, yet no cells are `NA`: the focal species
placed cleanly in every window, read entirely against the frame species it
co-occurs with. This is the out-of-frame projection idea, and it is the same
mechanism a new or held-out species uses to enter an already-fitted coordinate
system. The frame holds the coordinates fixed; a fresh focal species is read
against it without disturbing it, the way a held-out point is scored against a
model that was fitted without it. The one requirement is that the focal species
co-occur with at least one frame species in a window, since with no co-occurrence
there is no centroid to compute and that window's cell falls back to `NA`.

## Presence and cover weighting in trajectories

The `weights` argument changes what a co-occurrence is worth when the centroid is
formed. With `weights = "cover"`, the default when cover is present, each
co-occurrence is weighted by the geometric mean of the two covers, matching the
AbundPMI operator the frame itself uses, so an abundant associate pulls the
centroid harder than a sparse one. With `weights = "presence"`, every
co-occurring plot counts the same regardless of how much of either species was
present. The two readings differ whenever abundance carries signal, which is the
common case in vegetation data.

```{r weights-cover}
tr_cov <- species_trajectory(x, species = "focal", weights = "cover",
                            by = NULL, dim = 6, min_occurrence = 2)
tr_pre <- species_trajectory(x, species = "focal", weights = "presence",
                            by = NULL, dim = 6, min_occurrence = 2)
```

Reading the first frame coordinate under each weighting on one pair of axes shows
the tracks part company. They agree on the broad direction, because the
underlying turnover is the same, and they disagree on the detail, because cover
re-weights which associates dominate the centroid in each window.

```{r weights-plot}
dc <- as.data.frame(tr_cov); dp <- as.data.frame(tr_pre)
rng <- range(c(dc$d1, dp$d1))
plot(dc$center, dc$d1, type = "b", pch = 19, ylim = rng,
     xlab = "decade", ylab = "first frame coordinate (d1)")
lines(dp$center, dp$d1, type = "b", pch = 17, lty = 2)
legend("topright", c("cover", "presence"), pch = c(19, 17), lty = c(1, 2), bty = "n")
```

Cover is the sharper instrument when the abundance values mean something, since it
lets a dominant associate steer the centroid the way it steers the community.
Presence is the safer instrument when covers are ordinal guesses, patchily
recorded, or absent, since it asks only the question the data can answer, which
species were there together. The choice is a statement about how far the
abundance numbers can be trusted, and it rewards a deliberate decision over the
default.

## Community trajectories and novelty over time

`species_trajectory()` follows individual species. `community_trajectory()`
follows the plots themselves and asks how novel each window's communities are
relative to a reference. The species frame is fitted once on all plots, every
window's plots are pooled into that shared frame, and novelty is the per-plot
mean distance to the `k` nearest reference communities.

```{r comm}
ct <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5)
ct
```

The print lists the reference window (the first, 1990, by default) and the
mean novelty per window. The `novelty` table holds the full summary: window
label, center, plot count, and the mean and median per-plot novelty against the
reference.

```{r comm-novelty}
ct$novelty
```

Mean novelty climbs across the decades, because each later window's communities
have drifted further from the 1990 baseline as the `focal`-carrying plots take on
more of the `b` pool. The reference window sits at novelty zero up to the nearest-
neighbour averaging, since it is being compared to itself. The median sits beside
the mean in the table for a reason: novelty is a per-plot quantity, and a window
can carry a few highly novel plots against a stable majority. When the mean runs
well above the median, a minority of plots is doing the drifting; when the two
track each other, the whole window has shifted together. Reading both keeps a
single unusual plot from being mistaken for a community-wide change. The choice
of `k` sets how forgiving the novelty measure is: a small `k` flags a plot as
novel the moment it sits far from its very nearest reference neighbours, while a
larger `k` averages over more of the reference cloud and asks the plot to be far
from a broader baseline before it scores high.

```{r comm-plot}
nv <- ct$novelty
plot(nv$center, nv$mean_novelty, type = "b", pch = 19,
     xlab = "decade", ylab = "mean novelty vs reference",
     main = "community novelty relative to the 1990 baseline")
```

The reference is a choice, not a fixed point. Passing `reference =` a later window
label or index re-bases the novelty against that window instead of the first.
Setting the reference to the final decade measures how far each window sits from
the *end* state, which inverts the trend: early communities are now the novel
ones.

```{r comm-ref}
ct_end <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5,
                              reference = "2020")
ct_end$novelty
```

The reference can also be supplied as an index (`reference = 4` picks the fourth
window) or as an external `specvec_community` or matrix, which lets a window be
scored against a baseline that lives outside the time series entirely, such as a
reference community assembled from undisturbed sites.

```{r comm-ref-idx}
ct_idx <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5,
                              reference = 4)
identical(ct_idx$reference, "2020")
```

The `communities` field holds the per-window plot-by-dim matrices in the shared
frame, so the embeddings are available for similarity work, clustering, or
plotting beyond the novelty summary. It is a list with one matrix per window,
keyed by window label, each row a plot and each column a frame dimension.

```{r comm-communities}
lengths(lapply(ct$communities, nrow))
dim(ct$communities[["1990"]])
```

Because every matrix sits in the one frame, the window centroids are comparable
directly. Stacking the per-window mean position and tracking it across decades
gives a community-level analogue of the species track: the average plot moving
through the frame as its composition drifts. We read the first two frame
coordinates of each window centroid and trace them in order.

```{r comm-centroid}
cen <- t(sapply(ct$communities, colMeans))
plot(cen[, 1], cen[, 2], type = "b", pch = 19,
     xlab = "frame d1", ylab = "frame d2",
     main = "community centroid path through the frame")
text(cen[, 1], cen[, 2], rownames(cen), pos = 3)
```

The centroid walks across the frame in decade order, the community-scale version
of the same turnover the focal species traced one species at a time.

## Practical guidance

A few rules of thumb keep trajectories honest.

**Read the support before trusting a cell.** A focal species placed from two
co-occurring plots gives a noisier centroid than one placed from eighty. The
`$support` matrix on a species trajectory and the `$n_plots` column on both
`$windows` and `$novelty` report exactly how much data backs each cell. As a
working rule, treat a cell with support below 5 plots as indicative only and a
cell with support below 3 as unreliable, the same floor the default
`min_occurrence = 5` applies to the frame itself. The support matrix is small
enough to scan directly before any conclusion is drawn from a track.

```{r guide-support}
range(tr$support)
which(tr$support < 5, arr.ind = TRUE)
```

A track that looks dramatic only in a window where support has collapsed is a
track to distrust. The cheapest defense is to print `min(tr$support)` alongside
any displacement and refuse to read movement out of a cell the data cannot back.

**Match window width to the question.** Narrow windows resolve fine temporal
structure but spread the plots thin, raising the chance of `NA` cells and unstable
centroids. Wide windows pool more plots per cell and steady the estimate at the
cost of temporal resolution. The `by` break vector is the dial; widening it is the
first move when cells go sparse.

**Choose presence or cover deliberately.** With `weights = "cover"` (the default
when cover is present) co-occurrence is weighted by the geometric mean of covers,
matching AbundPMI, so abundant associates pull the centroid harder. With
`weights = "presence"` every co-occurring plot counts the same. Cover is the
sharper signal when the abundance is meaningful; presence is the safer choice when
covers are ordinal guesses or missing.

**Pick a reference window that means something.** Community novelty is always
relative to its reference. The first window is a sensible default for "how far
have we come", a final or undisturbed window suits "how far are we from a target",
and an external baseline lets a whole time series be scored against ground that
sits outside it.

**Reuse the frame across analyses.** Both verbs accept `frame_embedding =` a
pre-fitted `specvec_embedding`. Fitting one frame and passing it to both
`species_trajectory()` and `community_trajectory()` puts the species tracks and
the community novelty in the same coordinate system, so the two readings line up
exactly.

**When trajectories stop being reliable.** Two failure modes deserve a name.
The first is the empty cell: a focal species with no co-occurrence in a window
has no centroid, and the coordinate columns come back `NA`. This is honest
behaviour, the function declining to invent a position, and `as.data.frame(tr,
na.rm = TRUE)` drops those rows so the gap does not leak into a plot. The second
is the swinging cell: a window holding only a handful of plots gives a centroid
that lurches on individual records, because one unusual plot is a large fraction
of the pool. A track that is steady through well-populated windows and erratic
through a thin one is reporting the thinness of that window. The remedy for both
is the same, widen the window with the `by` break vector until each cell clears
the support floor, and accept the loss of temporal resolution that buys. A frame
fitted on too few species fails earlier and louder, since the default
`min_occurrence` will drop most species and the frame will refuse to fit below
two; that error is a signal to pool windows or lower the floor on purpose.

`integration_trajectory()` is the applied specialization of this machinery for
alien-integration questions: it places a neophyte with `species_trajectory()`,
places the native community per window in the same frame, and reads the distance
between them over time. The mechanics here are general; the alien case is worked
end to end, with ReSurvey anchoring and an EVA example, in
`vignette("specvec-integration")`.
