---
title: "Methods: from plots to embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods: from plots to embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(specvec)
```

This vignette defines the maths SpecVec runs. It is the theory companion to the
quickstart, which shows the verbs in use, and to the extending vignette, which
shows how to add a recipe of your own. Here we stay on the algorithm: what each
operator measures, how each factorization reads vectors out of it, and why the
two halves snap together the way they do. Every method follows one pipeline:

```
plot x species (+ cover)
        |
        v
   co-occurrence operator        (counts, PMI, cover-weighted PMI, chi-square, CLR)
        |
        v
   factorization                 (eigen, implicit SVD, or GloVe)
        |
        v
   species vectors  ->  community vectors (pooled)
```

A method is a registered `(weighting, factorization)` pair. `abund_pmi`, `pmi`,
`ca`, `glove`, and `clr` are points on this grid, and `specvec_methods()` lists
them.

```{r methods}
specvec_methods()
```

## Algorithm overview

The pipeline above has two moving parts and a fixed handshake between them. A
*weighting* turns the plot-by-species data into an operator: a square species
association, a raw count matrix, or a set of matrix-vector closures over the
contingency table. A *factorization* turns that operator into a species-by-dim
matrix of vectors. The handshake is the operator's `kind` field. Each weighting
stamps its output with one of three kinds, each factorization declares the one
kind it accepts, and the engine checks the two agree before any numerical work
begins. The three kinds and their consumers are fixed:

```
kind "sym"      -> symmetric species x species $M       consumed by  eigen
kind "counts"   -> species x species co-occurrence $M    consumed by  glove
kind "implicit" -> matvec closures $Af/$Atf + readout    consumed by  svd
```

The five shipped methods place themselves on this grid as follows. `abund_pmi`
and `pmi` and `clr` all build a `"sym"` operator and hand it to `eigen`; they
differ only in how the symmetric matrix is filled. `ca` builds an `"implicit"`
operator and hands it to `svd`. `glove` builds a `"counts"` operator and hands
it to its own objective factorizer. Adding a sixth method is one registration,
not a new code path, because dispatch reads the method record rather than
branching on a name. The extending vignette walks that registration step by
step; this vignette explains what the five built-in operators compute.

Two properties hold across all of them and are worth stating once. The fit is
deterministic: ids are stored sorted, and each embedding column is sign-fixed so
its largest-magnitude entry is positive, with no random component anywhere in
the matrix path, so two fits of the same data return identical matrices. And the
community readout is uniform: whatever recipe produced the species vectors, a
plot is placed at the weighted average of the vectors of the species it holds, so
plot vectors stay comparable across methods.

## Input and co-occurrence

`specvec()` reads a long table and builds two sparse matrices over a sorted
species index: presence `P` (plot by species, 0/1) and, when cover is supplied,
`COV` (plot by species, cover in `[0, 1]`). Duplicate plot-species rows are
aggregated by the `duplicates` rule (`"max"` by default), and ordinal cover
scales convert through `cover_scale`. The data vignette covers those readers; we
use a tiny hand-built table so the operator entries stay checkable by eye.

```{r data}
df <- data.frame(
  plot    = c("p1","p1","p2","p2","p3","p3","p4","p4"),
  species = c("A","B","A","B","A","C","B","C"),
  cover   = c(80, 20, 50, 50, 60, 40, 40, 60)
)
x <- specvec(df, "plot", "species", abundance = "cover")
x
```

The raw co-occurrence count of two species is the number of plots holding both.
For the presence matrix `P` that count is `crossprod(P)`: entry `[a, b]` sums the
plots where both `a` and `b` are present, and the diagonal `[a, a]` is the
occurrence count of `a`. `cooc_matrix()` returns this operator directly.

```{r cooc-counts}
Cc <- cooc_matrix(x, "counts", min_occurrence = 1)
as.matrix(Cc)
```

Reading the diagonal gives the occurrence counts: A is in three plots, B in
three, C in two. The off-diagonal `[A, B]` is 2, the two plots (p1, p2) where A
and B both occur. This counts matrix is the raw operand. Everything downstream is
a transform of it or a cover-weighted analogue of it.

## AbundPMI in full

AbundPMI is the default. It computes pointwise mutual information on a
cover-weighted co-occurrence operator rather than on raw counts. Writing
`W = sqrt(COV)` element-wise, the operator and its factorization are:

```
W       = sqrt(COV)                                element-wise, plot x species
A       = t(W) %*% W                               species x species
A[a, b] = sum_p sqrt(cover[p, a] * cover[p, b])    off-diagonal: geometric means
f[a]    = A[a, a] = sum_p cover[p, a]              diagonal: cover mass of a
PPMI[a, b] = max( log( A[a, b] * n_plots / (f[a] * f[b]) ), 0 )   for a != b
emb        = V_k %*% diag(sqrt(Lambda_k))          top-k eigenpairs of PPMI
```

Each step has a plain reading. `A = t(W) %*% W` is the cover-weighted
co-occurrence: a plot where both species occur contributes the geometric mean of
their two covers, so a pair of two dominant species weighs more than a pair where
one is a trace. The diagonal `f[a]` is the total cover of species `a` summed over
plots, the cover-weighted analogue of occurrence frequency. The PPMI transform
compares the observed association `A[a, b]` against what independent placement
with those marginals would give, `f[a] * f[b] / n_plots`, takes the log of the
ratio, and clips anything below zero. Clipping keeps the matrix sparse and
non-negative, and zeroing the diagonal removes a species' trivial association
with itself. The result is symmetric, so the factorization is an
eigendecomposition, and each eigenvector is scaled by the square root of its
eigenvalue so that dot products in the embedding approximate the PPMI entries.

The eigen step keeps only the positive eigenvalues. SpecVec asks `RSpectra` for
the `k` largest algebraic eigenpairs of the sparse operator, falls back to a
dense symmetric `eigen()` when the sparse solver returns nothing usable, takes
the eigenvectors whose eigenvalues clear a small positive threshold, and scales
each by the square root of its eigenvalue. When fewer than `dim` positive
eigenvalues exist, the remaining columns are filled with zeros so the matrix
always has the requested width.

```{r abundpmi}
emb <- species_embedding(x, method = "abund_pmi", dim = 2, min_occurrence = 1)
emb
species_similarity(emb, "A", "B")
```

### Verifying the formula

A theory vignette earns its keep by showing the formula and the code agree on a
case small enough to follow. We take the `[A, B]` entry of the cover-weighted
PPMI and compute it by hand from the cover values, then compare it against the
operator SpecVec builds through `cooc_matrix(x, "abundance_pmi")`. The cover
matrix on the `[0, 1]` scale is the percent values divided by 100.

```{r abundpmi-byhand-cov}
cov_AB <- rbind(p1 = c(A = 0.80, B = 0.20), p2 = c(A = 0.50, B = 0.50),
                p3 = c(A = 0.60, B = 0.00), p4 = c(A = 0.00, B = 0.40))
A_AB <- sum(sqrt(cov_AB[, "A"] * cov_AB[, "B"]))          # off-diagonal A[A,B]
f_A  <- sum(cov_AB[, "A"]); f_B <- sum(cov_AB[, "B"])     # cover masses
c(A_AB = A_AB, f_A = f_A, f_B = f_B)
```

Only p1 and p2 contribute to `A[A, B]`, the two plots where both occur, and each
contributes the geometric mean of the covers. The marginals `f_A` and `f_B` sum
each species' cover over all four plots. With `n_plots = 4` the PPMI entry is the
clipped log-ratio:

```{r abundpmi-byhand-ppmi}
n_plots  <- 4
ppmi_AB  <- max(log(A_AB * n_plots / (f_A * f_B)), 0)
ppmi_AB
```

Now the same entry from the package operator. `cooc_matrix(x, "abundance_pmi")`
returns the symmetric PPMI matrix over the kept species, and reading `[A, B]`
recovers the hand value.

```{r abundpmi-check}
M <- cooc_matrix(x, "abundance_pmi", min_occurrence = 1)
as.matrix(M)["A", "B"]
all.equal(as.numeric(as.matrix(M)["A", "B"]), ppmi_AB)
```

The two match, which confirms the operator the code builds is the operator the
formula describes. The same identity holds for every off-diagonal entry; we
checked one pair because one pair is enough to pin the formula.

## PMI against AbundPMI as an A/B

Plain PMI runs the identical path with `W = P`, the binary presence matrix. The
association becomes `A = t(P) %*% P`, the raw co-occurrence counts from the first
section, and the marginal `f[a]` becomes the occurrence count rather than the
cover mass. AbundPMI is exactly plain PMI with presence replaced by `sqrt(COV)`,
which is why abundance enters as a clean A/B comparison rather than a fork:
swap the operand, keep the transform and the factorization. The cover-free
operator is reachable through `cooc_matrix(x, "ppmi")`, and the two operators
differ on data that carries cover.

```{r ab-operators}
P_ppmi <- as.matrix(cooc_matrix(x, "ppmi", min_occurrence = 1))
A_ppmi <- as.matrix(cooc_matrix(x, "abundance_pmi", min_occurrence = 1))
round(rbind(presence = P_ppmi["A", ], cover = A_ppmi["A", ]), 3)
```

The two rows give species A's PPMI to every species under presence weighting and
under cover weighting. They diverge because the cover operand replaces a yes/no
co-occurrence with the geometric mean of the two covers, so a pair that shares
cover-heavy plots gets pulled together harder than a pair that merely shares the
plot list. Fitting both methods produces different embeddings from the same data.

```{r ab-fit}
emb_pmi <- species_embedding(x, method = "pmi",       dim = 2, min_occurrence = 1)
emb_abp <- species_embedding(x, method = "abund_pmi", dim = 2, min_occurrence = 1)
c(pmi = species_similarity(emb_pmi, "A", "B"),
  abund_pmi = species_similarity(emb_abp, "A", "B"))
```

When a dataset carries no cover column, `abund_pmi` has no `COV` to take the
square root of, so it falls back to presence PMI and records that on the object
with a message. The geometric-mean weighting is the whole of the abundance
treatment: with cover it sharpens the operator, without cover it is plain PMI.

## Factorization details

The eigen factorization is shared by the three `"sym"` methods. It decomposes the
symmetric operator, keeps the eigenpairs with positive eigenvalues, and scales
each eigenvector by the square root of its eigenvalue. The square-root scaling is
what makes the inner product of two species vectors approximate the
corresponding operator entry, since the operator equals `V Lambda V'` and
`(V Lambda^{1/2})(V Lambda^{1/2})' = V Lambda V'`. The solver path is
sparse-first with a dense fallback: SpecVec calls `RSpectra::eigs_sym` for the
top algebraic eigenvalues, and if that returns nothing finite and positive it
densifies the operator, symmetrizes it against numerical drift, and runs base
`eigen()`. Columns beyond the count of positive eigenvalues are zero-padded, so
the returned matrix always has the requested `dim` even when the operator has low
rank.

```{r eigen-dim}
emb6 <- species_embedding(x, method = "abund_pmi", dim = 6, min_occurrence = 1)
dim(emb6$V)
```

Here three kept species admit at most two positive eigenvalues, so columns three
through six are zeros padded to fill the requested width. The leading columns
carry the largest eigenvalues, so the first two dimensions hold the dominant
gradient and a projection onto them is a readable map.

Truncating to the top `dim` eigenpairs is a low-rank approximation of the
operator. The full PPMI matrix has as many positive eigenvalues as it has rank,
and keeping the largest ones retains the directions along which species
associations vary most. A continental dataset has thousands of species and a
PPMI operator of high rank, so the choice of `dim` trades fidelity against size:
a larger `dim` reconstructs the operator more closely and a smaller one keeps the
embedding compact. The eigenvalues decay, so the first few dozen dimensions
usually carry most of the structure, which is why a few dozen dimensions stand in
for thousands of raw species columns. The square-root scaling on each retained
eigenvector then weights a dimension by how much association it explains, so the
leading dimensions dominate any distance or cosine computed on the vectors.

The sign convention makes the result deterministic. Eigenvectors and singular
vectors are unique only up to sign, so SpecVec flips each embedding column to put
its largest-magnitude entry positive. The flip is magnitude-invariant, so dot
products and cosines are unchanged, and it uses no random number, so two fits of
the same data return byte-identical matrices.

```{r determinism}
e1 <- species_embedding(x, method = "abund_pmi", dim = 2, min_occurrence = 1)
e2 <- species_embedding(x, method = "abund_pmi", dim = 2, min_occurrence = 1)
identical(e1$V, e2$V)
max(abs(e1$V - e2$V))
```

## Correspondence analysis (ca)

`ca` is correspondence analysis: the singular value decomposition of the
chi-square residuals of the plot-by-species contingency table. The residual at
cell `[p, s]` is the observed count minus its independence expectation, divided
by the square root of the expectation, the standard chi-square standardization.
A dense version would form that residual matrix and take its SVD. SpecVec takes
the SVD without ever forming it. The weighting returns two closures, one applying
the residual matrix to a vector and one applying its transpose, built from the
row sums, column sums, and the sparse contingency. `RSpectra::svds` drives the
decomposition through those matrix-vector products, and a readout maps the right
singular vectors to species coordinates by dividing out the column scaling and
multiplying by the singular values. The operator carries `kind = "implicit"`,
which is what routes it to `svd` rather than `eigen`.

```{r ca}
emb_ca <- species_embedding(x, method = "ca", dim = 2, min_occurrence = 1)
emb_ca
nearest_species(emb_ca, "A", n = 2)
```

The implicit route is the reason `ca` scales: a continental contingency table
has tens of thousands of species columns, and densifying its residual matrix is
the step that would not fit in memory. Driving the SVD through closures keeps the
contingency sparse throughout. `ca` is the reference method the benchmark scores
the others against.

## GloVe (glove)

`glove` is the one objective-factorizer. Rather than decomposing a prepared
matrix, it minimizes a weighted least-squares loss directly on the raw
co-occurrence counts. Each species gets a vector and a bias, and the loss asks
the dot product of two species vectors plus their biases to match the log of
their co-occurrence count, with a weighting function that caps the influence of
the most common pairs at a cutoff `x_max`. The fit runs by stochastic gradient
descent, and the species vector is the sum of the two weight matrices the
optimizer learns. The operator GloVe consumes is `kind = "counts"`, the raw
co-occurrence matrix, and the loss and bias terms live inside the factorizer
rather than in a prepared operand.

GloVe needs the `text2vec` package, a Suggests dependency, for its solver. When
`text2vec` is absent and `method = "glove"` is forced, the fit stops with a clean
message pointing at the missing package; inside `compare_embeddings()` the method
is skipped with a note rather than failing the whole comparison. The chunk below
runs the fit only when `text2vec` is installed, so the vignette knits either way.

```{r glove, eval = requireNamespace("text2vec", quietly = TRUE)}
emb_glove <- species_embedding(x, method = "glove", dim = 2, min_occurrence = 1,
                               glove_iter = 50)
dim(emb_glove$V)
```

Because GloVe brings its own loss, it sits apart from the eigen and SVD methods:
those read vectors out of a fixed matrix, GloVe searches for vectors that fit an
objective. It is the registry's example of an objective factorization, which is
why its `kind` is `"objective"` and it accepts the raw counts operand directly.

## The compositional fork (clr)

`clr` takes the other route to abundance. Rather than weighting co-occurrence by
cover, it treats each plot as a composition and works in Aitchison geometry. The
species operator is the centered-log-ratio covariance, and it is estimated
without imputing zeros into the full plot-by-species table. The estimation runs
in two steps. First the variation matrix:

```
T[i, j] = Var_p( log( cover[p, i] / cover[p, j] ) )   over plots where both occur
```

`T[i, j]` is the variance of the log-ratio of the two species' covers, taken over
the plots where both are present. A pair seen together in fewer than two plots
has no defined log-ratio variance, and those cells are filled with the largest
observed variation, the compositionally most dissimilar value. Estimating each
pair on its own co-present plot set, rather than zero-filling the whole table, is
the approximation SparCC makes, and it is what lets the operator be built from
sparse data. Second the double-centering that turns the variation matrix into a
covariance:

```
G = -1/2 * J T J     with    J = I - 11'/S
```

This is the distance-to-Gram identity. Because `clr_i - clr_j =
log(cover_i / cover_j)`, the variation matrix satisfies `T[i, j] = G_ii + G_jj -
2 G_ij`, and double-centering inverts that relation to recover the CLR covariance
`G`. The centering is done as row-and-column-mean subtraction, an order
`S`-squared operation, so no dense matrix product is formed. The result is
symmetric and dense, and unlike a PPMI operator it can carry negative entries,
since a covariance has no non-negativity constraint. That genuinely different
operand is why `clr` registers through the capability machinery as its own
`"sym"` weighting rather than reusing the PMI path: it produces a symmetric
matrix `eigen` can take, filled from a compositional model of its own.

```{r clr}
emb_clr <- species_embedding(x, method = "clr", dim = 2, min_occurrence = 1)
emb_clr
nearest_species(emb_clr, "A", n = 2)
```

The two abundance routes were settled by benchmark. On the neutral co-occurrence
metrics the cover-weighted PMI operand recovers more held-out signal than the
compositional one at the sparsity of continental vegetation data: `abund_pmi`
leads `clr` on both held-out raw co-occurrence and link-prediction AUC, while
`clr` scores close to correspondence analysis, and habitat recovery is a
near-tie. `abund_pmi` stays the default, and `clr` is available for the
compositional treatment when a study wants Aitchison geometry. The benchmark
vignette reports the comparison; the exact numbers depend on the sample.

The compositional operator is reached through `species_embedding(method = "clr")`
rather than `cooc_matrix()`, whose `weighting` argument exposes the four
co-occurrence operators (`counts`, `ppmi`, `abundance_pmi`, `chi_square`). The
CLR covariance is built inside the `clr` method when an embedding is fit.

## Rare-species filter and determinism

Two controls shape the operator before it is built. `min_occurrence` drops
species seen in fewer than that many plots, since a species present once or twice
gives an unstable vector that adds noise without structure. The drop happens
before the operator is built, so every method sees the same surviving species and
the operator is square over them. `min_cooccurrence` prunes species pairs that
share fewer than that many plots, which thins the operator's off-diagonal where a
single shared plot would otherwise create a spurious edge. Both cutoffs are
recorded on the fitted object under `preprocessing`, so an embedding reports the
filtering that produced it.

```{r filter}
emb$preprocessing[c("min_occurrence", "min_cooccurrence",
                    "n_species_kept", "n_species_total")]
```

The reason a rare species hurts the estimate is in the PMI formula. The marginal
`f[a]` of a species seen once is tiny, so the ratio `A[a, b] * n_plots / (f[a] *
f[b])` is dominated by that small denominator, and a single shared plot can
inflate the log into a large positive PMI that reflects one accident of sampling.
Dropping species below `min_occurrence` removes those unstable rows before the
operator is formed, which steadies the spectrum the factorization reads. The
default of five plots is the value the benchmark settled on for continental data;
the tiny examples in this vignette lower it to one so every species survives.

Determinism follows from the sign convention shown earlier and from sorted ids.
Species are stored in sorted order, the operator is built in that order, and each
column is sign-fixed to put its largest-magnitude entry positive. None of these
steps draws a random number, so a refit reproduces the previous matrix exactly.
The one method with a stochastic solver is `glove`, whose SGD path is seeded by
`text2vec`; the matrix methods are deterministic by construction.

## Community pooling

A community embedding places each plot in the species space by pooling the
vectors of the species it contains. The readout is a weighted average:

```
U[p, ] = sum_s w[p, s] * V[s, ] / sum_s w[p, s]
```

With `weights = "cover"` (the default when cover is present) the weights are the
plot's cover values, so a dominant species steers the plot vector toward its own
corner of the space. With `weights = "presence"` the weights are 0/1, giving the
plain mean of the present species' vectors, where every present species gets an
equal vote regardless of how much ground it covers. Comparing the two placements
of a plot shows how much abundance, rather than the species list alone, is
shaping its position.

```{r community}
comm      <- community_embedding(x, embedding = emb, weights = "cover")
comm_pres <- community_embedding(x, embedding = emb, weights = "presence")
round(rbind(cover = comm$U["p1", ], presence = comm_pres$U["p1", ]), 3)
```

The readout is identical across methods. Whether the species vectors came from
`abund_pmi`, `ca`, `glove`, or `clr`, a plot is the weighted average of its
species, so plot vectors stay comparable no matter which recipe built the species
space underneath. Keeping the community step uniform is a deliberate choice: it
means a study can change the species recipe and still read plot similarities,
novelty, and trajectories the same way. The downstream community verbs,
`community_similarity()` and `community_novelty()`, operate on these plot
vectors, and the communities vignette covers them.

## Comparison to alternatives and the benchmark verdict

The operator-to-factorization view differs from running a single fixed
ordination. A classical analysis fixes one transform and one decomposition, for
instance the chi-square residual SVD that `ca` reproduces, and reports its
coordinates. SpecVec keeps the transform and the decomposition as separate,
swappable registry entries, so a modelling choice is a choice of operand and the
factorization follows from the operand's kind. The same eigen solver reads
vectors out of presence PMI, cover-weighted PMI, and the CLR covariance, because
all three present a symmetric matrix; correspondence analysis differs by
presenting an implicit operand instead. The grid view is what makes the abundance
question a clean A/B, the compositional route a registered sibling rather than a
rewrite, and a new idea a single registration.

`compare_embeddings()` scores the methods under one protocol and decides a winner
by a rule fixed before the results are seen. A method beats the reference (`ca`
by default) only if it exceeds it on both neutral co-occurrence metrics, held-out
raw Spearman and link-prediction AUC, by more than two pooled standard
deviations. Fixing the criterion in advance keeps the comparison honest: there is
no scanning for the metric on which a favoured method happens to look best, and a
method that ties, or wins one metric while tying the other, is reported as
`partial` or `no advantage` rather than promoted. The companion vignette
`vignette("specvec-benchmark")` runs it in full, including the optional `glove`
method. On continental vegetation data this verdict selects `abund_pmi`, and its
lead over correspondence analysis widens with the number of plots.
