---
title: "Methods: from plots to embeddings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methods: from plots to embeddings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(specvec)
```

This vignette defines the maths SpecVec runs. Every method follows one pipeline:

```
plot x species (+ cover)
        |
        v
   co-occurrence operator        (counts, PMI, cover-weighted PMI, or chi-square)
        |
        v
   factorization                 (eigen, implicit SVD, or GloVe)
        |
        v
   species vectors  ->  community vectors (pooled)
```

A method is a registered `(weighting, factorization)` pair. `ca`, `pmi`,
`abund_pmi`, and `glove` are points on this grid, and `specvec_methods()` lists
them.

## Input and co-occurrence

`specvec()` reads a long table and builds two sparse matrices over a sorted
species index: presence `P` (plot by species, 0/1) and, when cover is supplied,
`COV` (plot by species, cover in `[0, 1]`). Duplicate plot-species rows are
aggregated by the `duplicates` rule (`"max"` by default), and ordinal cover
scales convert through `cover_scale`.

```{r data}
df <- data.frame(
  plot    = c("p1","p1","p2","p2","p3","p3","p4","p4"),
  species = c("A","B","A","B","A","C","B","C"),
  cover   = c(80, 20, 50, 50, 60, 40, 40, 60)
)
x <- specvec(df, "plot", "species", abundance = "cover")
x
```

The raw co-occurrence count of two species is the number of plots holding both,
which is `crossprod(P)`. The diagonal is each species' occurrence count.

## AbundPMI, the default

AbundPMI is pointwise mutual information computed on a cover-weighted
co-occurrence operator. Writing `W = sqrt(COV)` element-wise:

```
A[a, b] = sum_p sqrt(cover[p, a] * cover[p, b])   # = t(W) %*% W, off-diagonal
f[a]    = A[a, a] = sum_p cover[p, a]             # cover mass of species a
PPMI[a, b] = max( log( A[a, b] * n_plots / (f[a] * f[b]) ), 0 )   for a != b
emb        = V_k %*% diag(sqrt(Lambda_k))         # top-k eigenpairs of PPMI
```

Each co-occurring plot contributes the geometric mean of the two covers, so a
pair of two abundant species weighs more than a pair where one is a trace. The
marginal `f[a]` is the total cover of species `a`, the cover-weighted analogue
of occurrence frequency. Only positive PMI is kept (PPMI), and the diagonal is
excluded. The operator is symmetric, so the factorization is an eigendecomposition
scaled by the square roots of the positive eigenvalues.

Plain PMI is the same path with `W = P`, the binary presence matrix. AbundPMI is
exactly PMI with presence replaced by `sqrt(COV)`, which makes the abundance
effect a clean A/B comparison.

```{r abundpmi}
emb <- species_embedding(x, method = "abund_pmi", dim = 2, min_occurrence = 1)
emb
species_similarity(emb, "A", "B")
```

When a dataset has no cover column, `abund_pmi` falls back to presence PMI and
records that on the object. The cover-weighted operator is also reachable
directly through `cooc_matrix(x, "abundance_pmi")`.

## Baselines

`ca` is correspondence analysis: the singular value decomposition of the
chi-square residuals of the plot-by-species contingency table. SpecVec applies it
through matrix-vector products, so the dense contingency table is never formed.
`pmi` is the presence version of AbundPMI above. `glove` is the GloVe
objective-factorizer from `text2vec` (an optional dependency); it brings its own
weighted least-squares loss and bias terms, and reads species vectors from the
sum of its two weight matrices.

```{r methods}
specvec_methods()
```

## Rare-species filter and determinism

`min_occurrence` drops species seen in fewer than that many plots before the
operator is built, since singletons give unstable vectors. `min_cooccurrence`
prunes rare species pairs. Both are recorded on the fitted object.

Eigen and SVD vectors are unique only up to sign. SpecVec fixes the sign of each
embedding column so its largest-magnitude entry is positive, which is
deterministic and leaves dot products and cosines unchanged. Two fits of the same
data return identical matrices.

## Community pooling

A community embedding places each plot in the species space by pooling the
vectors of the species it contains:

```
U[p, ] = sum_s w[p, s] * V[s, ] / sum_s w[p, s]
```

With `weights = "cover"` (the default when cover is present) the weights are
cover; with `weights = "presence"` they are 0/1, giving the plain mean of the
present species' vectors. `community_similarity()` and `community_novelty()`
operate on these plot vectors.

```{r community}
comm <- community_embedding(x, embedding = emb, weights = "cover")
comm
```

## The benchmark verdict

`compare_embeddings()` scores the methods under one protocol and decides a winner
by a rule fixed before the results are seen: a method beats the reference only if
it exceeds it on both neutral co-occurrence metrics (held-out raw Spearman and
link-prediction AUC) by more than two pooled standard deviations. The companion
vignette `vignette("specvec-benchmark")` runs it. On continental vegetation data
this verdict selects `abund_pmi`, and its lead over correspondence analysis grows
with the number of plots.
