---
title: "Tracking species and communities through time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracking species and communities through time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(specvec)
```

Vegetation data often arrives stamped with time: plots surveyed across decades,
a ReSurvey campaign returning to the same locations, a chronosequence read off
land-use history. The natural question is whether a species or a community has
moved through compositional space, and where it went. SpecVec answers this with
two front-door verbs, `species_trajectory()` and `community_trajectory()`, both
built on one idea: fit a single coordinate system once, then read every time
window out in that system.

## The cross-time alignment problem

The obvious approach is to embed each decade on its own. Fit one species
embedding on the 1990s plots, another on the 2010s plots, and compare the two
coordinate matrices. The approach fails for a numerical reason. An eigendecomposition
or singular value decomposition returns vectors
that are unique only up to sign, and when eigenvalues are close, up to rotation
within the degenerate subspace. SpecVec fixes each column's sign deterministically
(see `vignette("specvec-methods")`), so refitting the *same* data twice gives the
same matrix. Two *different* fits carry no such guarantee. The axes of the 1990s
embedding and the axes of the 2010s embedding are separate bases for separate
matrices. A species sitting at coordinate `(1.2, -0.4)` in one fit and
`(1.2, -0.4)` in the other has not necessarily stayed put; the two coordinate
pairs are written in different languages.

People patch this with Procrustes rotation: find the orthogonal transform that
best maps one fit onto the other, apply it, then compare. That works when the
two communities are mostly the same and only a few species moved. It breaks
exactly when the data are interesting. Procrustes assumes a shared structure to
align to, so if the species pool itself turned over between decades, the
alignment absorbs the very signal being measured, and a real compositional shift
gets rotated away. The rotation is also estimated from the data it then judges,
which is circular.

The fixed-frame solution sidesteps all of this. One *frame* embedding is fitted
once, on a stable reference set of species. Each time window is then projected
into that frame without refitting: a focal species is placed, per window, at the
cover-weighted centroid of the frame species it co-occurs with in that window.
The frame never moves, so there is nothing to align. A point in 1990 and a point
in 2010 are coordinates in the *same* basis, and their difference is a genuine
displacement. There is no per-window rotation, no Procrustes step, and no
circularity, because the thing measured (the focal species' position) is never
part of the thing that defines the axes (the frame).

The data that drives this needs a time column, supplied to `specvec()` at build
time. The simulation below codes four decades. A focal species rides one species
pool early and drifts toward a second pool late, while a backbone of unrelated
plots holds steady throughout.

```{r sim}
sim_time <- function(seed = 1) {
  set.seed(seed)
  a <- paste0("a", 1:8); b <- paste0("b", 1:8)
  decs <- c(1990, 2000, 2010, 2020)
  rows <- list(); pid <- 0L
  for (di in seq_along(decs)) {
    share <- (di - 1) / (length(decs) - 1)
    for (i in 1:100) {
      pid <- pid + 1L
      pool <- if (runif(1) < share) b else a
      sp <- c(sample(pool, sample(3:6, 1)), "focal")
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decs[di])
    }
    for (i in 1:50) {
      pid <- pid + 1L
      sp <- sample(a, sample(3:6, 1))
      rows[[pid]] <- data.frame(plot = paste0("p", pid), species = sp,
                                cover = round(runif(length(sp), 5, 90), 1),
                                decade = decs[di])
    }
  }
  do.call(rbind, rows)
}

x <- specvec(sim_time(), "plot", "species", abundance = "cover", time = "decade")
x
```

The `focal` species starts among the `a` pool and is pulled toward the `b` pool
as the decades pass. The `a`-only backbone plots keep the `a` species anchored,
so they make a sensible frame.

## A single window

A frame is itself just a species embedding. To see what one window looks like in
isolation, fit an embedding on a slice of the data with the `time` argument.
A numeric vector selects a closed interval, so `time = c(1990, 2000)` trains on
every plot whose decade falls between 1990 and 2000.

```{r window}
emb_early <- species_embedding(x, time = c(1990, 2000), dim = 6, min_occurrence = 2)
emb_early
```

This is a valid embedding of the early decades. The trouble is the second one.

```{r window2}
emb_late <- species_embedding(x, time = c(2010, 2020), dim = 6, min_occurrence = 2)
emb_late
```

`emb_early` and `emb_late` are two independent fits. Each has its own sign-fixed
axes, but those axes were chosen against different co-occurrence operators, so the
6 dimensions of one do not correspond to the 6 dimensions of the other. Comparing
`emb_early$V` to `emb_late$V` entry by entry measures the difference between two
coordinate systems as much as any difference between two decades. This is the
alignment problem in miniature, and it is the reason `species_trajectory()`
exists: it fits the frame once and reuses it for every window, so the comparison
is clean by construction.

## Species trajectories

`species_trajectory()` traces one or more focal species through the fixed frame.
The frame defaults to every species except the focal ones, so the focal species
move through a background they do not help define. Each window places the focal
species at the cover-weighted centroid of the frame species it co-occurs with in
that window, and a focal species never contributes to its own position.

```{r traj}
tr <- species_trajectory(x, species = "focal", by = NULL, dim = 6,
                         min_occurrence = 2)
tr
```

The print shows the frame (here the 16 non-focal species, embedded once), the
four windows, and the per-cell support range. With `by = NULL`, one window is cut
per distinct decade. The tidy form spreads the array into one row per
focal-species-by-window cell, carrying the window center, the support, and the
`dim` coordinate columns `d1` through `d6`.

```{r traj-df}
d <- as.data.frame(tr)
d
```

The `support` column is the number of window plots that actually contain the
focal species, and `center` is the numeric window center used for ordering and
plotting. The `windows` table records the same window metadata directly, with the
plot count per window.

```{r traj-windows}
tr$windows
```

A cell is `NA` in the coordinate columns when the focal species co-occurs with no
frame species in that window, so its centroid is undefined. That happens in
sparse windows, and `as.data.frame(tr, na.rm = TRUE)` drops those rows. Here the
focal species occurs in roughly 100 plots per decade, so every cell is backed by
ample co-occurrence and none are `NA`.

The displacement of the focal species between consecutive windows is the
Euclidean distance between successive rows of the coordinate matrix. The frame is
fixed, so this distance is a real measure of how far the species' associates
moved, not an artifact of refitting.

```{r traj-plot}
dcols <- grep("^d[0-9]+$", names(d))
co <- as.matrix(d[order(d$center), dcols])
step <- sqrt(rowSums((co[-1, ] - co[-nrow(co), , drop = FALSE])^2))
mids <- (d$center[order(d$center)][-1] + d$center[order(d$center)][-nrow(co)]) / 2

plot(mids, step, type = "b", pch = 19,
     xlab = "decade midpoint", ylab = "displacement in frame",
     main = "focal species: movement between successive decades")
```

The displacement is largest across the middle decades, where the simulated pull
from the `a` pool to the `b` pool is steepest, and smaller at the ends. The first
coordinate alone tells a similar story: tracking `d1` against the window center
shows the focal species sliding along the frame's leading axis as its co-occurrents
turn over.

```{r traj-d1}
plot(d$center, d$d1, type = "b", pch = 19,
     xlab = "decade", ylab = "first frame coordinate (d1)",
     main = "focal species: position on the leading frame axis")
```

Both views read the same fixed frame, so a movement on the plot is a movement in
the data.

## Choosing windows

The `by` argument controls how the time axis is split. The default `by = NULL`
makes one window per distinct time value, which suits decade-coded data where the
decades are the natural units.

```{r win-default}
tr$windows
```

Four decades, four windows, each centered on its decade and holding 150 plots
(100 carrying the focal species plus 50 backbone). Passing a numeric break vector
instead bins the time axis with `cut()`, using closed lower bounds. The breaks
`c(1985, 2005, 2025)` cut two bins, an early bin spanning 1985 to 2005 and a late
bin spanning 2005 to 2025.

```{r win-breaks}
tr2 <- species_trajectory(x, species = "focal", by = c(1985, 2005, 2025),
                          dim = 6, min_occurrence = 2)
tr2$windows
```

The window centers are now the midpoints of the break intervals, 1995 and 2015,
and each window pools two decades, so `n_plots` is 300 rather than 150. Coarser
windows hold more plots per window, which steadies the centroid in each cell while
lowering temporal resolution. Both readings describe the same trajectory at two
grains, and the break vector sets the grain.

## Community trajectories and novelty over time

`species_trajectory()` follows individual species. `community_trajectory()`
follows the plots themselves and asks how novel each window's communities are
relative to a reference. The species frame is fitted once on all plots, every
window's plots are pooled into that shared frame, and novelty is the per-plot
mean distance to the `k` nearest reference communities.

```{r comm}
ct <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5)
ct
```

The print lists the reference window (the first, 1990, by default) and the
mean novelty per window. The `novelty` table holds the full summary: window
label, center, plot count, and the mean and median per-plot novelty against the
reference.

```{r comm-novelty}
ct$novelty
```

Mean novelty climbs across the decades, because each later window's communities
have drifted further from the 1990 baseline as the `focal`-carrying plots take on
more of the `b` pool. The reference window sits at novelty zero up to the nearest-
neighbour averaging, since it is being compared to itself.

```{r comm-plot}
nv <- ct$novelty
plot(nv$center, nv$mean_novelty, type = "b", pch = 19,
     xlab = "decade", ylab = "mean novelty vs reference",
     main = "community novelty relative to the 1990 baseline")
```

The reference is a choice, not a fixed point. Passing `reference =` a later window
label or index re-bases the novelty against that window instead of the first.
Setting the reference to the final decade measures how far each window sits from
the *end* state, which inverts the trend: early communities are now the novel
ones.

```{r comm-ref}
ct_end <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5,
                              reference = "2020")
ct_end$novelty
```

The reference can also be supplied as an index (`reference = 4` picks the fourth
window) or as an external `specvec_community` or matrix, which lets a window be
scored against a baseline that lives outside the time series entirely, such as a
reference community assembled from undisturbed sites.

```{r comm-ref-idx}
ct_idx <- community_trajectory(x, dim = 6, min_occurrence = 2, k = 5,
                              reference = 4)
identical(ct_idx$reference, "2020")
```

The `communities` field holds the per-window plot-by-dim matrices in the shared
frame, so the embeddings are available for similarity work, clustering, or
plotting beyond the novelty summary.

## Practical guidance

A few rules of thumb keep trajectories honest.

**Read the support before trusting a cell.** A focal species placed from two
co-occurring plots gives a noisier centroid than one placed from eighty. The
`$support` matrix on a species trajectory and the `$n_plots` column on both
`$windows` and `$novelty` report exactly how much data backs each cell. Windows
where support drops into the low single digits deserve a skeptical eye.

**Match window width to the question.** Narrow windows resolve fine temporal
structure but spread the plots thin, raising the chance of `NA` cells and unstable
centroids. Wide windows pool more plots per cell and steady the estimate at the
cost of temporal resolution. The `by` break vector is the dial; widening it is the
first move when cells go sparse.

**Choose presence or cover deliberately.** With `weights = "cover"` (the default
when cover is present) co-occurrence is weighted by the geometric mean of covers,
matching AbundPMI, so abundant associates pull the centroid harder. With
`weights = "presence"` every co-occurring plot counts the same. Cover is the
sharper signal when the abundance is meaningful; presence is the safer choice when
covers are ordinal guesses or missing.

**Pick a reference window that means something.** Community novelty is always
relative to its reference. The first window is a sensible default for "how far
have we come", a final or undisturbed window suits "how far are we from a target",
and an external baseline lets a whole time series be scored against ground that
sits outside it.

**Reuse the frame across analyses.** Both verbs accept `frame_embedding =` a
pre-fitted `specvec_embedding`. Fitting one frame and passing it to both
`species_trajectory()` and `community_trajectory()` puts the species tracks and
the community novelty in the same coordinate system, so the two readings line up
exactly.

Trajectories stop being reliable when windows go sparse: a focal species with no
co-occurrence in a window yields an `NA` cell, and a window with only a handful of
plots gives a centroid that swings on individual records.

`integration_trajectory()` is the applied specialization of this machinery for
alien-integration questions: it places a neophyte with `species_trajectory()`,
places the native community per window in the same frame, and reads the distance
between them over time. The mechanics here are general; the alien case is worked
end to end, with ReSurvey anchoring and an EVA example, in
`vignette("specvec-integration")`.
