---
title: "Community embeddings and novelty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Community embeddings and novelty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(specvec)
```

A species embedding gives every species a vector. This vignette moves up one
level: it places each plot in that same space, compares plots to each other and
to a reference set, and scores how far a plot sits from familiar ground. The
worked examples run on a small simulated dataset where plots draw their species
from latent niches, so a plot's position in the embedding has a meaning we can
check against the way it was generated.

Three functions carry the work. `community_embedding()` pools species vectors
into one vector per plot. `community_similarity()` compares those plot vectors,
pairwise within one set or across two sets. `community_novelty()` scores each
plot by its distance to a reference body of communities. All three operate on the
same plot-by-dim matrix, so the path runs once from species vectors to plot
vectors and then branches into comparison and novelty without refitting.

The simulation builds `M` plots over a two-dimensional environment. Each species
has a niche centre, and a plot holds a species with probability that falls off
with the squared distance between the plot's environment and the species centre.
Cover is the relative niche match, scaled so the best-matched species in a plot
reads 100.

```{r sim}
sim_plots <- function(M = 300, S = 40, seed = 1) {
  set.seed(seed)
  mu <- matrix(rnorm(S * 2), S, 2)
  env <- matrix(rnorm(M * 2), M, 2)
  rows <- lapply(seq_len(M), function(p) {
    d2 <- rowSums((mu - matrix(env[p, ], S, 2, byrow = TRUE))^2)
    prob <- exp(-d2 / 2)
    present <- which(runif(S) < prob)
    if (length(present) < 2) present <- order(prob, decreasing = TRUE)[1:2]
    data.frame(plot = paste0("p", p),
               species = paste0("sp", present),
               cover = round(100 * prob[present] / max(prob[present]), 1))
  })
  do.call(rbind, rows)
}
```

## From species vectors to plot vectors

A species embedding `V` is a species-by-dim matrix: row `s` is the vector for
species `s`. A plot is a set of species, sometimes with cover. The community
embedding turns that set into a single point by pooling the vectors of the
species the plot contains, weighted by how much of each species is there:

```
U[p, ] = sum_s w[p, s] * V[s, ] / sum_s w[p, s]
```

The numerator adds up the species vectors, each scaled by its weight in the
plot; the denominator divides by the total weight, so `U[p, ]` is a weighted
average that lands inside the cloud of the plot's species. Two plots that share
most of their species, and weight them similarly, land near each other. A plot
made of niche specialists from one corner of the environment lands in that
corner of the embedding. The readout is the same regardless of which method
produced `V`, so plot vectors from different fits stay comparable as long as the
underlying species vectors agree.

The weighted average has a useful consequence. A plot vector is never an
arbitrary point; it is a convex combination of its species' vectors, so it sits
in the hull of the species it contains. A plot with one species sits on that
species' vector. A plot with many species spread across the embedding sits
near their weighted centroid, which tends toward the middle of the cloud. The
geometry of the species embedding therefore propagates upward: if two species
sit close because they co-occur, plots built from either of them sit close too.
Species that the embedding never saw, because they fell under `min_occurrence`,
carry no vector and drop out of the pooling, so the plot vector is built from the
species that survived the rare-species filter.

The build starts with `specvec()`, which reads the long table into presence and
cover matrices. A species embedding follows, and `community_embedding()` pools
it into plot vectors.

```{r build}
df <- sim_plots(M = 300, S = 40, seed = 1)
x <- specvec(df, "plot", "species", abundance = "cover")
emb <- species_embedding(x, method = "abund_pmi", dim = 8, min_occurrence = 3)
comm <- community_embedding(x, embedding = emb)
comm
```

The print line reports the number of plots, the embedding dimension, the pooling
rule, and the provenance of the species vectors. The plot matrix `U` carries one
row per plot and one column per embedding dimension.

```{r dims}
dim(comm$U)
head(rownames(comm$U), 4)
```

With 300 plots and `dim = 8`, `U` is 300 by 8. Each row is a plot's coordinates
in the eight-dimensional species space, ready for the similarity and novelty
tools below.

## Cover vs presence pooling

The weight `w[p, s]` is where cover enters. With `weights = "cover"` the weight
is the plot's cover for species `s`, so a species covering 80 of a plot pulls the
plot vector four times as hard as one covering 20. With `weights = "presence"`
every present species carries weight 1, and the plot vector is the plain mean of
the vectors of the species on the list. The two share the same species
embedding; they differ only in how they average it. The same `emb` flows into
both calls below, so any difference in the resulting plot positions comes from
the weighting rule alone, which makes the comparison a clean A/B test of cover
against presence.

```{r pooling}
comm_cov <- community_embedding(x, embedding = emb, weights = "cover")
comm_pre <- community_embedding(x, embedding = emb, weights = "presence")
comm_cov$pooling
comm_pre$pooling
```

The two plot matrices are built from the same `V` and the same species lists,
yet they place the plots differently. A quick way to see this is the per-plot
distance between the cover vector and the presence vector: zero would mean the
two agree, and most plots sit well above zero.

```{r pooling-diff}
delta <- sqrt(rowSums((comm_cov$U - comm_pre$U)^2))
summary(delta)
```

The gap is largest in plots with one dominant species and several trace
companions. Cover pooling pulls such a plot toward the dominant's vector;
presence pooling treats the dominant and the traces alike and sits closer to the
centroid of the whole species list. In a plot where one species covers 90 and
two others cover 5 each, cover pooling weights the dominant eighteen times as
heavily as either companion, so the plot vector lands almost on the dominant's
position. Presence pooling weights all three equally and lands a third of the way
between them. The two readings of the same plot answer different questions about
what the plot is.

The choice follows the question. Cover pooling suits abundance-driven analyses,
where a plot carpeted by one species should read as that species' kind of place,
and where the cover values come from a consistent scale. Presence pooling suits
species-list work, where the data are checklists with no reliable abundance, or
where every recorded species should count once regardless of how much of it was
there. When the input has no cover column, `community_embedding()` falls back to
presence pooling on its own and reports that it did so, so a presence-only
dataset behaves sensibly without any extra argument.

The scatter below shows the plots in the first two community dimensions under
cover pooling, with four plots labelled. Position carries the latent niche: plots
near each other share their abundant species.

```{r pooling-plot, fig.width = 7, fig.height = 5}
set.seed(2)
lab <- sample(rownames(comm_cov$U), 4)
plot(comm_cov$U[, 1], comm_cov$U[, 2],
     pch = 16, col = "grey60", cex = 0.7,
     xlab = "community dim 1", ylab = "community dim 2",
     main = "Plots in community space (cover pooling)")
points(comm_cov$U[lab, 1], comm_cov$U[lab, 2], pch = 16, col = "black")
text(comm_cov$U[lab, 1], comm_cov$U[lab, 2], labels = lab, pos = 3, cex = 0.8)
```

## Geometry of pooled plot vectors

Pooling is a weighted average of species vectors, and the averaging fixes where a
plot can land. A weighted average with non-negative weights is a convex
combination, so a plot vector always sits inside the convex hull of the species
it contains. The plot cannot land outside the cloud of its own species. A plot
built from three species sits somewhere in the triangle their vectors span; add a
fourth species and the plot can move into the tetrahedron, never beyond it.

Two limiting cases make the geometry concrete. A plot holding a single species
sits on that species' vector under presence pooling, and a plot holding two
species at equal weight sits at their midpoint. We can check both. To keep every
embedding species available for the lookup, the demo data carries one extra plot
listing all species, then reads the geometry of the two plots of interest.

```{r geometry-check}
sp <- emb$species[1:2]
demo <- rbind(
  data.frame(plot = "solo", species = sp[1], cover = 50),
  data.frame(plot = "pair", species = sp, cover = c(50, 50)),
  data.frame(plot = "all",  species = emb$species, cover = 50))
xg <- specvec(demo, "plot", "species", abundance = "cover")
cg <- community_embedding(xg, embedding = emb, weights = "presence")
c(solo = max(abs(cg$U["solo", ] - emb$V[sp[1], ])),
  pair = max(abs(cg$U["pair", ] - (emb$V[sp[1], ] + emb$V[sp[2], ]) / 2)))
```

Both differences are zero: the one-species plot lands on its species' vector, and
the two-species plot lands at the midpoint of the two vectors. The exact landing
holds for presence pooling, where the weights are ones and sum to the number of
species. Cover pooling keeps the direction but can shorten the vector, because the
pooling floors the total weight at one to keep a faint plot from blowing up, so a
single-species plot whose total cover weight is below one sits along the species'
vector at reduced length rather than exactly on it.

The hull property guides interpretation. A plot near the centre of the embedding
cloud is a generalist mixture, drawing on species from across the space, so its
vector averages out toward the middle. A plot near a species corner is dominated
by that species and its close neighbours, so its vector sits out at the edge. Two
plots near the same corner share that dominant kind of species; two plots near
the centre may share nothing specific and still sit close, because both are broad
mixtures. The novelty section below uses this directly: a plot that pools to a
thinly populated region between corners is one whose species rarely keep company,
and that is the geometric signature of an unusual community.

## Fitting fresh vs reusing an embedding

`community_embedding()` can fit a species embedding on the spot. Passing
`method`, `dim`, and the rare-species filters without an `embedding` argument
runs `species_embedding()` internally and pools the result.

```{r fit-fresh}
comm_fresh <- community_embedding(x, method = "abund_pmi", dim = 8,
                                  min_occurrence = 3)
comm_fresh
```

The alternative is to fit once and pass the fitted object through `embedding =`.
When the internal fit uses the same method, dimension, and filters as a pre-fit
embedding, the two paths produce the same species vectors and therefore the same
plot matrix `U`.

```{r fit-reuse}
comm_reuse <- community_embedding(x, embedding = emb)
max(abs(comm_fresh$U - comm_reuse$U))
```

The maximum element-wise difference is zero to numerical precision: the fresh fit
and the reused embedding describe the same plots in the same coordinates. The
equality holds because the species embedding is deterministic. SpecVec fixes the
sign of every embedding column, so two fits of the same data with the same
filters return byte-identical species vectors, and pooling them gives byte
-identical plot vectors. Reusing one embedding has a practical payoff. Several
community sets pooled from one frame share an axis system, so their plot vectors
compare directly and the fit runs once instead of once per set. The reuse path
also lets the species embedding come from a larger or more authoritative dataset
than the plots being pooled, which is the usual setup when a small survey is
placed in the coordinate system of a continental reference.

`normalize = TRUE` rescales each plot vector to unit length after pooling, which
sends every plot to the surface of a sphere and makes cosine comparisons depend
on direction alone. Leave it off when distances between plots should reflect both
direction and length; turn it on when only the compositional direction matters.

```{r normalize}
comm_norm <- community_embedding(x, embedding = emb, normalize = TRUE)
comm_norm$normalized
range(sqrt(rowSums(comm_norm$U^2)))
```

## Placing new plots in an existing frame

The reuse path supports the workflow that matters most in practice: fit a frame
once on a large reference body of plots, then drop a separate, smaller set of
plots into that same frame and compare them against the reference. The frame is
the species embedding. As long as every new plot is pooled through it, the new
plot vectors land in the reference coordinate system and the comparison tools
read them on equal terms with the reference plots.

We split the simulated plots into a reference block and a query block, fit the
species embedding on the reference alone, and then pool both blocks through that
one embedding.

```{r place-fit}
all_plots <- rownames(comm_cov$U)
ref_block <- all_plots[1:240]
qry_block <- all_plots[241:300]
x_ref <- specvec(df[df$plot %in% ref_block, ], "plot", "species", abundance = "cover")
emb_ref <- species_embedding(x_ref, method = "abund_pmi", dim = 8, min_occurrence = 3)
comm_ref <- community_embedding(x_ref, embedding = emb_ref)
```

The query block never entered the fit. Its plots are placed afterwards by pooling
their species through `emb_ref`, the embedding learned from the reference. The
result is a community object whose rows are the query plots and whose columns are
the reference frame's dimensions.

```{r place-pool}
x_qry <- specvec(df[df$plot %in% qry_block, ], "plot", "species", abundance = "cover")
comm_qry <- community_embedding(x_qry, embedding = emb_ref)
dim(comm_ref$U)
dim(comm_qry$U)
```

The query matrix has 60 rows and the same 8 columns as the reference, so the two
sit in one space. `community_similarity()` with the reference passed through
`reference =` returns a 60-by-240 matrix: each query plot scored against every
reference plot. Reading off the column of the largest entry per row gives each
query plot its closest match in the reference.

```{r place-match}
S_qr <- community_similarity(comm_qry, reference = comm_ref, metric = "cosine")
dim(S_qr)
best_match <- apply(S_qr, 1, function(row) colnames(S_qr)[which.max(row)])
head(best_match, 3)
```

`community_novelty()` then scores how far each query plot sits from the reference
body, against the same reference coordinates.

```{r place-novelty}
nov_qry <- community_novelty(comm_qry, comm_ref, k = 5)
round(range(nov_qry), 3)
```

This is the realistic version of placing a small survey in a continental frame.
The reference embedding stands in for a large, authoritative dataset; the query
plots are a fresh survey too small to fit a stable embedding on its own. Pooling
the survey through the reference embedding gives every new plot coordinates that
mean the same thing as the reference coordinates, so similarity and novelty read
correctly without refitting. One condition makes the placement work: every
species in the query data must be a species the reference embedding knows. The
pooling looks up each embedding species in the query data, so a query species
absent from the reference frame has no vector to contribute, and a query frame
missing reference species cannot be indexed. In practice a query is harmonised to
the reference taxonomy first, and species the reference never saw are dropped from
the query before pooling, the same way the rare-species filter drops them at fit
time.

## Community similarity

`community_similarity()` compares plot vectors. Called on one community object it
returns the self-similarity matrix: every plot against every other plot. With
`metric = "cosine"` the entries are cosine similarities in `[-1, 1]`, where 1 is
two plots pointing the same way in species space and a value near zero is two
plots with little in common.

```{r sim-self}
S <- community_similarity(comm_cov, metric = "cosine")
dim(S)
round(S[1:4, 1:4], 2)
```

The diagonal is 1, each plot is identical to itself. An off-diagonal cell close
to 1 marks a pair of plots whose abundant species overlap and point the same
direction; a cell near zero or below marks a pair drawing on different niches. In
the submatrix above, the larger cells flag plot pairs that share their dominant
species. Cosine reads composition through the embedding rather than through raw
species overlap, so two plots can score high even with no species in common, as
long as their species occupy nearby parts of the embedding. That is the
difference from a Jaccard or Bray-Curtis index computed on the species lists:
the embedding lets ecologically similar species stand in for each other, while a
set-overlap index counts only the species that match by name.

## Embedding similarity vs raw set overlap

The claim that the embedding sees structure a set-overlap index misses is worth
testing directly. We can build a Jaccard dissimilarity from the presence matrix
`x$P` in a few lines, with no extra package: Jaccard is one minus the size of the
shared species set divided by the size of the combined set. The presence matrix
carries plots in its rows and species in its columns, so a row of `x$P` greater
than zero is a plot's species list as a logical vector.

```{r jaccard-fn}
plots_sub <- rownames(comm_cov$U)[1:60]
P_sub <- as.matrix(x$P[plots_sub, , drop = FALSE]) > 0
jaccard <- function(a, b) {
  u <- sum(a | b)
  if (u == 0) 1 else 1 - sum(a & b) / u
}
```

Filling a 60-by-60 Jaccard matrix is a double loop over the plot pairs. Alongside
it we take the cosine similarity from the embedding and turn it into a
dissimilarity by subtracting from one, so both matrices read the same direction:
small means similar, large means different.

```{r jaccard-build}
n <- length(plots_sub)
Jac <- matrix(0, n, n, dimnames = list(plots_sub, plots_sub))
for (i in seq_len(n)) for (j in seq_len(n)) Jac[i, j] <- jaccard(P_sub[i, ], P_sub[j, ])
cos_dis <- 1 - community_similarity(comm_cov, metric = "cosine")[plots_sub, plots_sub]
ut <- upper.tri(Jac)
cor(Jac[ut], cos_dis[ut], method = "spearman")
```

The Spearman correlation between the two dissimilarities runs about 0.82 on this
set. The embedding and the species lists agree on the broad ordering of plot
pairs, which is the reassuring part: plots that share many species mostly do sit
close in the embedding. The 0.82 also leaves room for disagreement, and the
disagreement is where the embedding earns its place. The clearest case is a pair
of plots that share no species at all, which Jaccard scores as maximally
dissimilar by construction, yet which the embedding places close together.

```{r jaccard-pair}
zero_overlap <- Jac == 1
diag(zero_overlap) <- FALSE
emb_cos <- community_similarity(comm_cov, metric = "cosine")[plots_sub, plots_sub]
emb_cos[!zero_overlap] <- -Inf
hit <- which(emb_cos == max(emb_cos), arr.ind = TRUE)[1, ]
pa <- plots_sub[hit[1]]; pb <- plots_sub[hit[2]]
c(shared_species = sum(P_sub[hit[1], ] & P_sub[hit[2], ]),
  jaccard = Jac[hit[1], hit[2]],
  cosine = round(emb_cos[hit[1], hit[2]], 3))
```

The two plots `r pa` and `r pb` hold no species in common, so
their Jaccard dissimilarity is the maximum 1, and a checklist comparison files
them as unrelated. Their cosine similarity in the embedding is around 0.95. The
species that fill the two plots are different by name but sit in the same region
of the embedding, because they co-occur with a common third set of species across
the wider dataset, so the pooled plot vectors point almost the same way. A
set-overlap index cannot reach this conclusion: it has only the species names and
no notion that two unshared species play similar ecological roles. The embedding
carries that notion in the geometry, which is the reason to pool through it rather
than score the raw lists.

The disagreement runs the other way too. Two plots can share a species or two yet
sit apart in the embedding when their other species pull them into different
regions. Jaccard counts the shared name while the embedding weighs the whole
composition, so the two indices answer different questions. The Jaccard number
measures how much the species lists overlap, while the cosine measures how close
the communities sit in the learned species space. For ordination, gradient analysis, and novelty
scoring, the second question is usually the one of interest, which is why the rest
of this vignette works in the embedding.

A reference set goes in through the `reference` argument. The result has one row
per plot in `object` and one column per plot in `reference`, so it answers "how
does each of these plots compare with each of those". Splitting the plots into
two halves gives a worked cross-comparison.

```{r sim-ref}
plots_all <- rownames(comm_cov$U)
half <- length(plots_all) %/% 2
xa <- specvec(df[df$plot %in% plots_all[1:half], ], "plot", "species",
              abundance = "cover")
xb <- specvec(df[df$plot %in% plots_all[(half + 1):length(plots_all)], ],
              "plot", "species", abundance = "cover")
ca <- community_embedding(xa, embedding = emb)
cb <- community_embedding(xb, embedding = emb)
Scross <- community_similarity(ca, reference = cb, metric = "cosine")
dim(Scross)
```

`metric = "euclidean"` switches the readout to straight-line distance between
plot vectors. Distance grows as plots diverge, so it reads as a dissimilarity:
small values are near neighbours, large values are far apart. Cosine ignores
vector length and cares about direction; Euclidean distance feels both, and the
two rank plot pairs differently when plots vary in how strongly their species
load on the embedding. A species-poor plot and a species-rich plot built from
the same niche point the same way and score high on cosine, yet sit a real
distance apart in Euclidean terms because their vectors have different lengths.
Pairing `metric = "euclidean"` with `normalize = TRUE` at the pooling step
removes the length difference, so the distance then tracks direction alone and
the two metrics agree on ordering. The Euclidean distance from
`community_similarity()` is the same distance `community_novelty()` averages over
its nearest neighbours, so the two functions read plots through one geometry.

```{r sim-euclid}
D <- community_similarity(comm_cov, metric = "euclidean")
round(D[1:4, 1:4], 2)
```

### How far the two metrics agree

We can measure how much the choice of metric changes the ranking of plot pairs.
Take the cosine dissimilarity and the Euclidean distance over the whole community
set, and correlate them by rank across all plot pairs. A rank correlation of one
would mean the two metrics order every pair the same way.

```{r metric-rank}
cos_d <- 1 - community_similarity(comm_cov, metric = "cosine")
euc_d <- community_similarity(comm_cov, metric = "euclidean")
ut <- upper.tri(cos_d)
cor(cos_d[ut], euc_d[ut], method = "spearman")
```

The correlation is about 0.99 on this set. The two metrics mostly agree, because
the plot vectors here vary only modestly in length: the niche simulation gives
every plot a comparable number of species, so the pooled vectors sit at similar
distances from the origin. The agreement is not perfect, and the gaps show up in
the neighbour rankings of individual plots. We take the plot with the longest
vector and list its nearest neighbours under each metric.

```{r metric-neighbours}
len <- sqrt(rowSums(comm_cov$U^2))
foc <- names(which.max(len))
cos_nn <- names(sort(community_similarity(comm_cov)[foc, ], decreasing = TRUE))[2:4]
euc_nn <- names(sort(community_similarity(comm_cov, metric = "euclidean")[foc, ]))[2:4]
rbind(cosine = cos_nn, euclidean = euc_nn)
```

The two neighbour lists share their first two entries and diverge at the third.
Cosine and Euclidean agree on the very closest plots and start to disagree
further out, where a longer plot vector that points almost the same way scores
high on cosine yet sits a measurable distance away. Setting `normalize = TRUE` at
the pooling step erases that length difference, so the Euclidean distance then
depends on direction alone. The two metrics then rank every pair identically.

```{r metric-norm}
euc_dn <- community_similarity(comm_norm, metric = "euclidean")
cor(cos_d[ut], euc_dn[ut], method = "spearman")
```

The rank correlation between cosine dissimilarity and normalized Euclidean
distance is exactly one. After normalization the Euclidean distance between two
unit vectors is a monotone function of their cosine, so the two readouts carry the
same ordering. The practical rule that follows: if a downstream method works in
Euclidean distance but you want it to read composition the way cosine does, pool
with `normalize = TRUE` and the distance will behave like cosine without any
further adjustment.

An `image()` of a small cosine block shows the structure at a glance. Brighter
cells are more similar plot pairs; the bright diagonal is each plot against
itself.

```{r sim-image, fig.width = 5.5, fig.height = 5}
sub <- S[1:20, 1:20]
image(seq_len(20), seq_len(20), sub,
      col = hcl.colors(12, "YlOrRd", rev = TRUE),
      xlab = "plot", ylab = "plot",
      main = "Cosine similarity (first 20 plots)")
```

## Community novelty

Novelty asks a different question. It measures how far a plot sits from a body of
known communities. `community_novelty()` scores each plot in `object` by its mean
Euclidean distance to the `k` nearest plots in `reference`.
A plot whose `k` nearest reference neighbours are all close scores low; a plot
that sits far from everything in the reference set scores high. The score is a
distance, so it has no upper bound and is read relative to the spread of the
reference.

The mean-of-nearest-neighbours definition matters. A plot's novelty is the
average distance to its `k` closest reference plots. That choice sits between two
cruder options: distance to the single closest reference plot, which reacts to
one chance neighbour, and distance to the reference centroid, which ignores local
structure. Averaging over a handful of neighbours keeps the score steady against
a lone reference plot that happens to sit nearby, so a query plot with one close
neighbour and four distant ones still scores as novel. The score reads as
distance to a local pocket of the reference, which is the property that lets it
separate plots in dense regions from plots in sparse ones.

The setup splits the plots into a reference set and a query set, both pooled from
the same embedding so their coordinates align. The reference is the bulk of the
simulated plots; the query mixes ordinary plots with a handful built to land in
an unusual part of the species space.

```{r novelty-setup}
ref_plots <- plots_all[1:250]
qry_plots <- plots_all[251:300]
x_ref <- specvec(df[df$plot %in% ref_plots, ], "plot", "species",
                 abundance = "cover")
comm_ref <- community_embedding(x_ref, embedding = emb)
```

To force genuine novelty, build a few odd plots that pair the species sitting
farthest apart in the embedding. Such species rarely co-occur in the reference,
so a plot holding both pools to a point between the reference clusters, in a
thinly populated region.

```{r novelty-odd}
dmat <- as.matrix(dist(emb$V))
ut <- which(upper.tri(dmat), arr.ind = TRUE)
ord <- order(dmat[ut], decreasing = TRUE)
set.seed(7)
odd <- do.call(rbind, lapply(1:6, function(i) {
  pr <- ut[ord[i], ]
  data.frame(plot = paste0("odd", i), species = emb$species[pr],
             cover = round(runif(2, 40, 100), 1))
}))
qry_df <- rbind(df[df$plot %in% qry_plots, ], odd)
x_qry <- specvec(qry_df, "plot", "species", abundance = "cover")
comm_qry <- community_embedding(x_qry, embedding = emb)
```

With reference and query in hand, `community_novelty()` returns one score per
query plot. `k = 5` averages over the five nearest reference communities, which
smooths over a single unusually close or far neighbour while still tracking the
local density of the reference.

```{r novelty-run}
nov <- community_novelty(comm_qry, comm_ref, k = 5)
round(sort(nov, decreasing = TRUE)[1:8], 3)
```

The six odd plots take the top six places. They pair species the reference rarely
holds together, so their plot vectors land in a thinly populated region and their
nearest five reference neighbours sit far away, around 0.5 to 0.7 in embedding
distance. Ordinary query plots, built the same way as the reference, score below
0.3 because they fall inside the reference cloud where neighbours are close.

```{r novelty-check}
is_odd <- grepl("^odd", names(nov))
c(odd = mean(nov[is_odd]), ordinary = mean(nov[!is_odd]))
```

The mean novelty of the deliberately odd plots runs about four to five times the
ordinary query plots, which is the behaviour the score is built to give. A
histogram shows the bulk of plots at low novelty with a tail of outliers; the odd
plots sit in that tail, marked by the red rug.

```{r novelty-hist, fig.width = 7, fig.height = 4}
hist(nov, breaks = 12, col = "grey80", border = "white",
     xlab = "novelty (mean distance to 5 nearest reference plots)",
     main = "Distribution of community novelty")
rug(nov[is_odd], col = "red", lwd = 2)
```

### Three definitions of distance to a reference

The mean-of-k-nearest score is one of three natural ways to measure how far a
plot sits from a reference. We can compute all three for a single odd plot and
compare. The distance to the single nearest reference plot reacts to one
neighbour; the distance to the reference centroid averages over the whole
reference and ignores local structure; the mean of the five nearest sits in
between.

```{r novelty-defs}
R <- comm_ref$U
q <- comm_qry$U[which(is_odd)[1], ]
d_all <- sqrt(rowSums((R - matrix(q, nrow(R), ncol(R), byrow = TRUE))^2))
c(nearest = min(d_all),
  mean_k5 = mean(sort(d_all)[1:5]),
  centroid = sqrt(sum((q - colMeans(R))^2)))
```

The distance to the single nearest plot is around 0.48, the mean of the five
nearest around 0.52, and the distance to the centroid around 0.78. The centroid
distance reads largest because the reference fans out around its mean, so even a
plot inside the cloud sits some way from the average point. The centroid figure
also misses the local picture: it cannot tell a plot in a dense reference pocket
from one in a sparse pocket the same distance from the mean. The single-nearest
figure has the opposite weakness, swinging on one chance neighbour. The mean of
the five nearest reads the local pocket, which is the property
`community_novelty()` is built on.

### How k moves the scores

The value of `k` shifts the scores in a predictable way. We sweep `k` and record
the mean novelty of the odd plots, the mean for the ordinary query plots, and
their ratio.

```{r novelty-ksweep}
ks <- c(1, 3, 5, 10, 20)
sweep <- t(sapply(ks, function(k) {
  nv <- community_novelty(comm_qry, comm_ref, k = k)
  c(k = k, odd = mean(nv[is_odd]), ordinary = mean(nv[!is_odd]),
    ratio = mean(nv[is_odd]) / mean(nv[!is_odd]))
}))
round(sweep, 3)
```

Every score rises with `k`, because reaching for more neighbours pulls in plots
that sit further away. The ordinary plots climb faster than the odd ones, so the
ratio falls from near 7 at `k = 1` to about 3 at `k = 20`. The odd plots stay
flagged across the whole range, which is the steadiness the mean buys, yet a large
`k` blunts the contrast by averaging the reference's sparse pockets into its dense
ones. A moderate `k` keeps the odd plots far above the ordinary ones while still
smoothing single-neighbour noise.

### A plain matrix as the reference

`community_novelty()` reads either a `specvec_community` or a plain numeric matrix
of community vectors, so a fixed reference frame can be saved once and reused
across query batches.

```{r novelty-matrix}
R_mat <- comm_ref$U
nov_mat <- community_novelty(comm_qry, R_mat, k = 5)
identical(unname(nov_mat), unname(nov))
```

The scores from the matrix reference match the scores from the community object
exactly, because the function works on the underlying plot-by-dim matrix either
way. In a production setting the reference matrix is computed once on the
authoritative dataset and stored, and each incoming batch of plots is scored
against the stored matrix without rebuilding a community object, which keeps the
reference fixed across every batch.

The choice of `k` trades sensitivity against stability. Small `k` (1 or 2) reacts
to the single closest reference plot and flags anything off the beaten track, at
the cost of noise. Larger `k` averages over a neighbourhood and reports distance
to a region rather than a point, which is steadier when the reference is dense.
The function caps `k` at the number of reference communities, so passing a `k`
larger than the reference simply uses every reference plot. The reference may be
a `specvec_community` or a plain numeric matrix of community vectors, so a fixed
set of reference coordinates, saved once, can be reused across many query
batches without rebuilding a community object each time.

One detail keeps the comparison honest: the query and the reference must live in
the same coordinate system. Pooling both through the same species embedding `emb`
guarantees that. A novelty score computed against a reference pooled from a
different fit would mix two axis systems and report distances that mean nothing,
which is the same reason the similarity examples above pooled their two halves
from one shared `emb`.

## Practical guidance

A few rules of thumb cover most uses.

Choose pooling by what the data mean. Use `weights = "cover"` when abundance is
reliable and a dominant species should dominate the plot vector, so a stand
carpeted by one species reads as that species' kind of place. Use
`weights = "presence"` for checklists, for occurrence-only surveys, and whenever
every recorded species should count once. The default falls back to presence on
its own when the data carry no cover column.

Normalize when only direction matters. Leaving `normalize = FALSE` keeps both the
direction and the length of each plot vector, so a strongly loaded plot reads as
more extreme than a faint one. Set `normalize = TRUE` when plots should be
compared on composition alone, which is the common choice ahead of cosine
clustering, since every plot then sits on the unit sphere and length drops out.
Pairing `normalize = TRUE` with `metric = "euclidean"` makes the distance match
the cosine ordering, which is the setting to use when a downstream method expects
Euclidean input yet you want it to read composition.

Match the metric to the question. Cosine reads the compositional direction of a
plot and ignores how strongly its species load; Euclidean distance feels both
direction and length. For comparing what kind of community two plots are, cosine
is the usual choice. For novelty, the score is built on Euclidean distance, so
plots that load weakly on the embedding read as nearer the origin and can score
lower; normalize the pooling first if length should not enter the novelty score.

Place a small survey in a large frame by reusing the embedding. Fit the species
embedding on the large reference dataset, then pool the small survey through it
with `embedding =`. The survey plots land in the reference coordinate system and
compare directly to the reference with `community_similarity()` and
`community_novelty()`. Harmonise the survey's species to the reference taxonomy
first, since a survey species the reference never saw has no vector to contribute
and a survey missing reference species cannot be pooled through the frame.

Pool several community sets from one shared embedding. Fitting the species
embedding once and passing it through `embedding =` to each
`community_embedding()` call guarantees the sets share an axis system. Plot
vectors from different sets then compare directly, and the fit cost is paid once.
Refitting per set risks sign or scaling drift that breaks cross-set comparison.

Scale `k` to the reference. With a small reference set a large `k` averages over a
big fraction of all reference plots and washes out the local structure novelty is
meant to read; `community_novelty()` already caps `k` at the number of reference
communities. With a few hundred reference plots, `k` of 5 to 10 balances
sensitivity against noise. Keep `k` well below the reference size so the score
still reflects a local neighbourhood.

Give novelty a meaningful reference. The score is a distance to whatever
reference you pass, so a reference that is itself narrow makes ordinary plots look
novel, and a reference that already contains the odd communities makes them look
ordinary. A reference assembled to represent "what is known" gives a score that
reads as "how far from the known". This is also the limit to keep in mind:
novelty is relative to the reference, not an absolute outlier score, so a plot
flagged as novel against one reference can be unremarkable against another.
