---
title: "Getting started with specvec"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with specvec}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
```

```{r setup}
library(specvec)
```

## Introduction

specvec learns abundance-aware vectors for species and for whole communities from
ecological plot data. The input is the long table that nearly every vegetation
survey already produces: one row per plot-species record, optionally carrying a
cover value. From that table specvec places every species at a point in a shared
coordinate space, where species that keep the same company across plots land near
each other. Two species that repeatedly share habitat end up with similar vectors;
two species that never co-occur sit far apart.

Cover enters the geometry through the weighting, not as an afterthought. In the
default method, a co-occurring plot contributes the geometric mean of the two
species' covers, so a plot where both species dominate counts for more than a plot
where one is a trace. A species that is consistently abundant in the company it
keeps gets pulled harder toward that company than a species that merely brushes
past it. This is what "abundance-aware" means here: the same word ("co-occurrence")
gets a quantitative weight instead of a yes/no.

The second idea is that communities are first-class citizens of the same space.
Once species have vectors, a plot becomes the cover-weighted average of the
vectors of the species it holds. Plots therefore live in the same coordinate
system as species, and the distance between two plots reflects how similar their
weighted species composition is in embedding terms. That single readout supports
similarity matrices between plots, novelty scores for a plot against a reference
set, and trajectories over time. A plot vector and a species vector can sit side by
side and be compared with the same cosine, because they were built from the same
coordinates.

It helps to contrast this with the table that usually starts a vegetation
analysis. A plot-by-species matrix records who is present and how much, but it says
nothing directly about which species behave alike: that has to be inferred through
an ordination or a distance measure layered on top. An embedding bakes the
similarity into the coordinates themselves. Two species sit close exactly when
their cover-weighted co-occurrence is higher than chance would predict, so reading
off neighbours, clustering, and projecting all reduce to ordinary geometry in a
modest number of dimensions, typically a few dozen rather than the thousands of
raw species columns.

This vignette walks the full path on small simulated data: build a `specvec_data`
object, fit a species embedding, read off neighbours and pairwise similarities,
pool plots into community vectors, compare cover-weighted against presence-weighted
pooling, score novelty against a reference block, and run the shipped method
bake-off. Every step uses base R graphics so the example knits anywhere, and the
data are small enough (a few hundred plots) that each fit returns in a moment. The
closing section is a set of short, runnable pointers into the deeper vignettes.

## A worked dataset

A clean way to see the geometry work is to simulate it from a known cause. Give
each of `S` species a latent niche, a point `mu` in a two-dimensional environment.
Give each of `M` plots an environmental position `env`. A species occurs in a plot
with probability that falls off with the squared distance between the species'
niche and the plot's environment, and where it does occur its cover scales with
that same probability. Species with nearby niches will tend to show up together in
plots whose environment sits between them, which is the co-occurrence structure
specvec is built to recover.

The simulation is deliberately transparent so the embedding has a known answer to
match. Two species whose niches `mu` are close share many of the same plots and
both reach high cover there, which is exactly the cover-weighted co-occurrence the
default operator rewards; two species whose niches sit far apart rarely meet. If
the method works, the species map it produces should mirror the latent niche map
that generated the data, up to rotation and scale.

```{r sim}
sim_plots <- function(M = 300, S = 40, seed = 1) {
  set.seed(seed)
  mu  <- matrix(rnorm(S * 2), S, 2)
  env <- matrix(rnorm(M * 2), M, 2)
  rows <- lapply(seq_len(M), function(p) {
    d2 <- rowSums((mu - matrix(env[p, ], S, 2, byrow = TRUE))^2)
    prob <- exp(-d2 / 2)
    present <- which(runif(S) < prob)
    if (length(present) < 2) present <- order(prob, decreasing = TRUE)[1:2]
    data.frame(plot = paste0("p", p), species = paste0("sp", present),
               cover = round(100 * prob[present] / max(prob[present]), 1))
  })
  df <- do.call(rbind, rows)
  attr(df, "mu") <- mu
  df
}
df <- sim_plots()
head(df, 4)
```

The result is a long `plot, species, cover` table, exactly the shape `specvec()`
expects. The default run draws 300 plots over 40 species, and the falloff term
`exp(-d2 / 2)` keeps each plot to the handful of species whose niches lie near its
environment. A short guard promotes the two highest-probability species into any
plot that would otherwise hold fewer than two, so no plot is empty.

We pass the column roles by name and read cover on the percent scale (values run
from near 0 up to 100). Internally `specvec()` builds two sparse matrices over a
sorted species index: a presence matrix `P` of zeros and ones, and a cover matrix
`COV` rescaled into `[0, 1]`. The print method reports how many plots and species
survived, whether cover is present, and how duplicate plot-species rows were
aggregated; the `duplicates = "max"` default keeps the largest cover when a species
appears twice in one plot.

```{r build}
x <- specvec(df, plot = "plot", species = "species", abundance = "cover",
             cover_scale = "percent")
x
```

The object exposes its pieces directly. `x$species` and `x$plots` are the sorted id
maps, `x$P` and `x$COV` are the sparse matrices, and `x$meta` records the choices
above. The latent niches `mu` stay attached to `df` as an attribute, so after
fitting we can check whether species that were assigned nearby niches really did
end up as embedding neighbours, which is the property the whole method is meant to
recover.

## Species embedding

`species_embedding()` turns the data object into species vectors. The `method`
argument is shorthand for a registered `(weighting, factorization)` pair; the
default `"abund_pmi"` is abundance-weighted pointwise mutual information. It builds
a cover-weighted co-occurrence operator, keeps only the positive PMI entries, and
takes the top eigenvectors scaled by the square roots of their eigenvalues. The
practical effect is that two species sharing a lot of cover-heavy plots get a
large positive PMI and are pulled together, while pairs that co-occur no more than
chance contribute nothing. When a dataset carries no cover column, `abund_pmi`
falls back to the presence version and records that on the object.

Pointwise mutual information measures how much more two species co-occur than they
would if they were placed in plots independently. Raw counts favour common species,
which dominate any pair they touch simply by being everywhere; PMI divides the
observed co-occurrence by the product of the two species' marginals, so the score
reflects association rather than abundance alone. The "abund" part replaces the
binary presence with the square root of cover, which means a plot contributes the
geometric mean of the two covers to the pair's score, and each species' marginal
becomes its total cover mass instead of its plot count. Negative PMI values, the
pairs that co-occur less than chance, are clipped to zero before the
factorization, which keeps the operator sparse and the geometry stable.

```{r fit-species}
emb <- species_embedding(x, method = "abund_pmi", dim = 8, min_occurrence = 3)
emb
dim(emb$V)
```

`emb$V` is a species-by-dimension matrix with species as row names; here it is
eight-dimensional and holds one row per species that occurred in at least three
plots. `min_occurrence = 3` drops the rarest species before the operator is built,
since species seen once or twice give unstable vectors, and `emb$preprocessing`
records how many were kept against the total. The full list of fitted species is in
`emb$species`. The fit is deterministic: eigenvectors are unique only up to sign,
so specvec fixes each column's sign by its largest-magnitude entry, and two fits of
the same data return identical matrices.

The sticky readout is "which species sit next to this one". `nearest_species()`
ranks the closest species to a focal species by cosine similarity (the default) or
Euclidean distance, returning a tidy data frame.

```{r nearest}
focal <- emb$species[1]
nearest_species(emb, focal, n = 5)
```

`species_similarity()` answers the pairwise question directly. Pass two species and
it returns one cosine in `[-1, 1]`; pass one species and it returns the full named
vector of similarities to every species, handy for sorting or thresholding.

```{r similarity}
others <- emb$species[2:3]
species_similarity(emb, focal, others[1])
sim_all <- species_similarity(emb, focal)
round(sort(sim_all, decreasing = TRUE)[1:4], 3)
```

The cosine ranges over `[-1, 1]`: values near 1 mark species the embedding treats
as habitat partners, values near 0 mark species with no shared signal, and negative
values mark species that avoid each other. The named-vector form is the basis for
any downstream sort or threshold, for instance pulling every species within a
chosen similarity of a focal taxon.

The first two embedding columns carry the leading eigenvalues, so a scatter of
column 1 against column 2 gives a readable map of the dominant gradient. The full
geometry lives in all eight dimensions, and this projection flattens the rest, but
it is enough to see clusters form. We label a handful of species so the picture
stays uncluttered.

```{r plot-species, fig.alt = "Two-dimensional scatter of the first two species-embedding columns for a handful of species, each point labelled with its species id."}
V <- emb$V
show <- emb$species[1:12]
plot(V[show, 1], V[show, 2], pch = 19, col = "#3366aa",
     xlab = "embedding dim 1", ylab = "embedding dim 2",
     main = "Species in the first two embedding dimensions")
text(V[show, 1], V[show, 2], labels = show, pos = 3, cex = 0.7)
```

Points that cluster on this map are species the embedding judges to share habitat,
which traces back to the latent niches that generated the data.

## Community embeddings

Species vectors are the building block; community vectors are the payoff.
`community_embedding()` places each plot in the same space by pooling the vectors
of the species the plot contains. Passing the already-fitted embedding through
`embedding =` reuses those exact vectors instead of refitting, which keeps the
species and community spaces aligned and saves the fit. With `weights = "cover"`
(the default when cover is present) the pooling is a cover-weighted average, so a
dominant species steers the plot vector more than a trace; with
`weights = "presence"` it is the plain mean of the present species' vectors.

```{r fit-community}
comm <- community_embedding(x, embedding = emb, weights = "cover")
comm
dim(comm$U)
```

`comm$U` is a plot-by-dimension matrix with plots as row names, one row per plot
that contributed species in the fitted set. The pooled-vector readout is uniform
across embedding methods, so plot vectors stay comparable no matter which method
produced the species vectors underneath; the species space can change recipe while
the community readout stays a weighted average over it. A plot lands at the
weighted centre of its species, so a plot dominated by one cover-heavy species sits
close to that species' vector, and a plot spread evenly across several sits near
their average.

The presence-weighted version answers a slightly different question. It asks which
species are there at all rather than how much of each there is, and the two can
disagree for plots dominated by one species. Cover pooling lets a single abundant
species pull the plot toward its own corner of the space; presence pooling gives
every present species an equal vote regardless of how much ground it covers.
Comparing the two placements is a quick way to see how much abundance, rather than
mere species list, is shaping a given plot's position.

```{r presence-community}
comm_pres <- community_embedding(x, embedding = emb, weights = "presence")
comm_pres$pooling
```

`community_similarity()` returns a plot-by-plot matrix of cosines. Restricting it
to a few plots makes the structure easy to read: values near 1 mark plots with
near-identical weighted composition, values near 0 mark plots with little shared
signal.

```{r community-sim}
plots5 <- comm$plots[1:5]
sub <- structure(list(U = comm$U[plots5, , drop = FALSE], pooling = comm$pooling,
                      normalized = comm$normalized, from = comm$from, plots = plots5),
                 class = "specvec_community")
round(community_similarity(sub, metric = "cosine"), 2)
```

`community_novelty()` scores how unusual a plot is against a reference set: for
each scored plot it averages the distance to its `k` nearest reference
communities, so a plot that sits far from everything in the reference scores high.
We split the plots into a reference block and a query block and score the queries
against the reference.

```{r novelty}
ref_plots <- comm$plots[1:250]
qry_plots <- comm$plots[251:length(comm$plots)]
ref <- structure(list(U = comm$U[ref_plots, , drop = FALSE], pooling = comm$pooling,
                      normalized = comm$normalized, from = comm$from, plots = ref_plots),
                 class = "specvec_community")
qry <- structure(list(U = comm$U[qry_plots, , drop = FALSE], pooling = comm$pooling,
                      normalized = comm$normalized, from = comm$from, plots = qry_plots),
                 class = "specvec_community")
nov <- community_novelty(qry, reference = ref, k = 5)
round(sort(nov, decreasing = TRUE)[1:4], 3)
```

The plots at the top of that list are the ones whose weighted composition is least
like anything in the reference block. Averaging over the `k` nearest references
rather than the single closest one keeps the score from hinging on one quirky
reference plot; raising `k` smooths it further and lowers the scores of plots that
sit near a small reference cluster. On real data the reference is often a baseline
period or a reference region, and high-novelty plots flag compositions worth a
closer look, whether that is a community drifting away from its historical state or
a survey reaching habitat the reference never sampled.

## Comparing methods

specvec registers several embedding recipes, and `compare_embeddings()` scores them
side by side under one protocol so the choice is grounded in held-out performance
rather than taste. It filters the species set once, splits plots into train and
test, fits each method on the training plots through the same engine path, and
scores the methods on neutral tasks. Here we compare correspondence analysis
(`ca`), presence PMI (`pmi`), and the abundance-weighted default (`abund_pmi`) on
three co-occurrence metrics across two seeds.

```{r compare}
bench <- compare_embeddings(x, methods = c("ca", "pmi", "abund_pmi"),
                            metrics = c("cooc_ppmi", "cooc_raw", "link_auc"),
                            dim = 8, seeds = 1:2, min_occurrence = 3)
bench
```

The protocol matters as much as the scores. The species set is filtered once,
globally, before any split, so every method sees the same species and the
comparison is not confounded by one method quietly dropping rarer taxa. The plots
are then split into train and test, each method is fit on the training plots
through the same engine path that `species_embedding()` uses, and the held-out
plots supply the scoring signal. Each seed is an independent split, and the table
reports the across-seed mean and standard deviation so a method that wins by luck
on one split is visible as a wide spread.

The metrics read as follows. `cooc_raw` is the Spearman correlation between
embedding-space similarity and held-out raw co-occurrence counts: it asks whether
the geometry recovers how often pairs actually appear together in plots the model
never saw. `cooc_ppmi` is the same idea scored against held-out positive PMI rather
than raw counts, which downweights the common species that dominate raw counts.
`link_auc` is link-prediction AUC: given pairs that do co-occur and never-seen
negatives that do not, it measures how well embedding similarity separates them,
with 0.5 being a coin flip and 1.0 perfect separation. The three together cover
both the strength of association the geometry encodes and its ability to call a
pair present or absent.

The verdict line applies a rule fixed before the numbers are seen: a method beats
the reference (`ca` by default) only if it exceeds it on both neutral metrics
(`cooc_raw` and `link_auc`) by more than two pooled standard deviations. Fixing the
criterion in advance keeps the comparison honest: there is no scanning for the
metric on which a favoured method happens to look best. The same criterion is
reported whatever the outcome, so a method that merely ties, or wins on one metric
while tying the other, is labelled `partial` or `no advantage` rather than
promoted. With only two seeds on a few hundred simulated plots the standard
deviations here are wide and a clean two-sigma separation is unlikely; the verdict
sharpens as seeds and plots grow. On continental vegetation data this protocol
selects `abund_pmi`, and its lead over correspondence analysis widens with the
number of plots. The companion benchmark vignette runs the full comparison,
including the optional `glove` method, which factorizes the same co-occurrence
operator through a weighted least-squares objective and activates when the
`text2vec` package is installed.

## Where to go next

The five steps above are the whole core loop: build a data object, fit species
vectors, pool them into community vectors, score novelty, compare methods. Each
downstream topic has a dedicated vignette that picks up where this one stops. The
taste tests below are runnable sketches, each a few lines plus a pointer to where
the detail lives.

**Data prep and cover scales.** `specvec()` reads percent cover by default, and
also converts proportions and ordinal Braun-Blanquet codes, with a `duplicates`
rule for repeated plot-species rows.

```{r taste-data}
bb <- data.frame(plot = c("a","a","b","b"), species = c("X","Y","X","Z"),
                 bb = c("3", "+", "5", "2"))
xb <- specvec(bb, "plot", "species", abundance = "bb",
              cover_scale = "braun_blanquet", duplicates = "max")
xb$meta$cover_scale
```

The cover-scale lookups, custom `cover_mapping`, and aggregation rules are covered
in `vignette("specvec-data")`.

**Community novelty in depth.** Novelty against a reference set scales to large
reference blocks, and the interpretation shifts with the choice of reference and
`k`.

```{r taste-novelty, eval = FALSE}
ce  <- community_embedding(x, embedding = emb)
ref <- community_embedding(x_baseline, embedding = emb)
nov <- community_novelty(ce, reference = ref, k = 10)
hist(nov, main = "novelty against baseline")
```

How to choose a reference and read the novelty scores is covered in
`vignette("specvec-communities")`.

**Tracking change over time.** Store a plot-level time column on the data object and
specvec can fit per-window embeddings placed in one shared frame, then trace how a
species or a community moves across windows.

```{r taste-time, eval = FALSE}
xt <- specvec(df_time, "plot", "species", abundance = "cover", time = "decade")
traj <- species_trajectory(xt, species = "sp1")
plot(traj)
```

Windowed fits and trajectories for species and communities are in
`vignette("specvec-temporal")`.

**The maths, the bake-off, and extending it.** The operator-to-factorization
pipeline, including the compositional `clr` route, is in
`vignette("specvec-methods")`.

```{r taste-methods}
specvec_methods()
```

The full method comparison and the fixed verdict are in
`vignette("specvec-benchmark")`, and adding a new weighting or factorization to the
registry is in `vignette("specvec-extending")`.
