---
title: "Preparing data and cover scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data and cover scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(specvec)
```

Everything in SpecVec starts from one table and one call to `specvec()`. This
vignette covers the table's shape, the object the call returns, the cover scales
that turn an abundance column into a proportion, the rules for duplicated
records, the optional time and label columns, and what happens to invalid or
missing values. The companion vignette `vignette("specvec-methods")` picks up
from the object and defines the embedding maths.

The one call does the formatting work that vegetation analyses usually spread
across several preprocessing steps: it sorts the species and plots into a stable
index, builds the sparse presence and cover matrices the engine reads, converts
whatever cover scale the source used, and collapses any repeated records. The
sections that follow each take one of those steps and show its inputs, its
outputs, and the argument that controls it. The aim is that by the end you can
look at a vegetation table from any source and know which arguments to set.

## Expected input

`specvec()` reads a long table: one row per plot-species record. A plot column
identifies the site, a species column names the taxon, and an optional abundance
column carries cover. The same plot id repeats across its species, and the same
species id repeats across the plots that hold it. This is the shape most
vegetation databases export, where each row is one observation of one taxon in
one relevé, so in practice the export usually needs no reshaping before it
reaches `specvec()`.

The long form is what the co-occurrence operator wants. A wide site-by-species
table would need pivoting back to pairs of co-occurring species, and the long
table already lists them. Three plots and four species fit on the page; the
European examples later in this vignette run to hundreds of thousands of rows
with the same three columns.

```{r input}
df <- data.frame(
  plot    = c("p1", "p1", "p2", "p2", "p3", "p3"),
  species = c("Festuca", "Trifolium", "Festuca", "Bromus", "Festuca", "Carex"),
  cover   = c(40, 10, 60, 5, 30, 80)
)
df
```

Column roles are passed by name, so the columns can be called anything. Here
`plot` is the site id, `species` is the taxon, and `cover` is the abundance,
read on the default percent scale.

```{r input-build}
x <- specvec(df, plot = "plot", species = "species", abundance = "cover")
x
```

The first two arguments after `data` are the plot and species column names, and
both are required. Passing them by name keeps the call self-documenting and lets
the columns carry whatever names the source gave them, `PlotObservationID` and
`taxon` as readily as `plot` and `species`. The abundance column is optional:
without it, SpecVec builds presence only and the methods that need cover fall
back to presence. The `time` and `labels` arguments are optional too, covered
further down. A `specvec_data` object prints a short summary of its plots, species,
and settings, and the sections below open it up.

## What specvec() builds

The returned object is a list with named components. `P` and `COV` are the two
sparse matrices the engine reads; the rest are index maps and metadata.

```{r build-names}
names(x)
x$species
dim(x$P)
x$meta
```

`x$species` is the sorted vector of unique species ids, and `x$plots` is the
sorted vector of unique plot ids. These define the row and column order of every
matrix on the object, so a species' column position is the same in `P`, in
`COV`, and in the embedding the engine returns. Sorting fixes that order up
front, which is what makes two runs on the same data return identical matrices.
`dim(x$P)` is plots by species: 3 plots and 4 species in this example.

`P` is the presence matrix, plot by species, with a 1 where a species occurs in
a plot and an implicit 0 elsewhere. `COV` is the cover matrix, plot by species,
with each entry a cover proportion in `[0, 1]`. `COV` is `NULL` when no abundance
column is supplied. Both are stored sparse, so a plot that holds 30 of 50,000
species stores 30 entries, not 50,000, and an absent species costs nothing in
memory. The presence methods read `P`; the cover-weighted methods read `COV`;
the same object serves both.

```{r build-slice}
as.matrix(x$P)
round(as.matrix(x$COV), 2)
```

The percent covers from `df` divide by 100: a cover of 40 becomes 0.40, 80
becomes 0.80. `x$meta` records the settings used, including the duplicate rule,
the cover scale, the number of records read (`n_obs`), the number of duplicated
plot-species pairs found (`n_duplicates`), and whether an abundance column was
present (`has_abundance`). These fields are worth a glance after every build:
they confirm how many records survived validation and how the cover was read,
which is exactly the kind of input check that catches a wrong-scale or
wrong-column mistake before it reaches the embedding.

## Cover scales

The `cover_scale` argument tells SpecVec how to read the abundance column. Three
scales ship: `"percent"`, `"proportion"`, and `"braun_blanquet"`. Each maps the
raw values to a cover proportion in `[0, 1]`, which is the common currency the
cover-weighted methods expect, so the choice of scale is a reading instruction
rather than a transformation you apply yourself. `cover_from_scale()` exposes
that conversion on its own, which is handy for checking how a column will be read
before committing it to a build.

`"percent"` is the default and divides by 100, with values clamped to the
`[0, 100]` range first, so a stray 120 lands at 1.0 rather than overshooting.
`"proportion"` treats the column as already on `[0, 1]` and clamps to that range.
The difference between the two is one factor of 100, and picking the wrong one is
the most common cover mistake, which is why the standalone converter is the quick
way to see the values before they go in.

```{r scale-numeric}
cover_from_scale(c(0, 50, 100), scale = "percent")
cover_from_scale(c(0, 0.5, 1), scale = "proportion")
```

`"braun_blanquet"` reads ordinal cover-abundance codes and replaces each with
the percent midpoint of its cover class, then divides by 100 like the percent
scale. The codes are matched case-insensitively after trimming whitespace, so
`"R"` and `" r "` both read as `r`. The shipped codes are
`r`, `+`, `1`, `2`, `2m`, `2a`, `2b`, `3`, `4`, and `5`, with `2m`/`2a`/`2b` the
split forms of class 2. The midpoints run from 0.1 percent for `r` up to 87.5
percent for `5`, which is the centre of the 75 to 100 percent class.

```{r scale-bb}
cover_from_scale(c("r", "+", "2", "5"), scale = "braun_blanquet")
```

The Braun-Blanquet midpoints are a documented convention rather than a fixed
standard: authors differ on the exact class boundaries, and the split classes
`2m`, `2a`, and `2b` are not universal. SpecVec ships one widely used set of
midpoints and makes them overridable so your analysis can match whichever
convention your data follow. The `cover_mapping` argument replaces the lookup
with your own named numeric vector of percent midpoints. The names are the codes,
the values are the percents, and the same vector works through
`cover_from_scale()` or through `specvec()`. Here is a van der Maarel 1-9 ordinal
scale:

```{r scale-custom}
vdm <- c("1" = 0.1, "2" = 0.5, "3" = 3, "4" = 10, "5" = 18,
         "6" = 37.5, "7" = 62.5, "8" = 87.5, "9" = 97.5)
cover_from_scale(c("1", "5", "9"), scale = "braun_blanquet", mapping = vdm)
```

Passed to `specvec()`, the same mapping drives the cover column:

```{r scale-custom-build}
dv <- data.frame(
  plot = c("q1", "q1", "q2"),
  taxon = c("Festuca", "Carex", "Festuca"),
  vdm = c("9", "5", "1")
)
xv <- specvec(dv, "plot", "taxon", abundance = "vdm",
              cover_scale = "braun_blanquet", cover_mapping = vdm)
round(as.matrix(xv$COV), 3)
```

Any percent-midpoint scale, Domin or Londo or another, fits the same way: a
named numeric vector handed to `cover_mapping`. The mechanism is one lookup, so a
new scale is a new vector, not new code. A code absent from the mapping reads as
`NA` and is filled downstream, covered in the validation section below, so a
mapping that omits a rarely used class still builds rather than failing.

## Duplicate plot-species rows

A long table can carry the same plot-species pair more than once: two cover
estimates for one taxon in one plot, a taxonomy harmonization that merged two
names into one, or a join that left duplicates. The presence matrix `P` is
unaffected, since a species either occurs in a plot or it does not, but the cover
matrix needs a single value per pair. The `duplicates` argument decides how those
rows collapse into one cover. The default is `"max"`, which suits the merged-name
case where the largest cover is the safest single representative.

```{r dup-data}
dd <- data.frame(
  plot    = c("p1", "p1", "p1", "p2"),
  species = c("Festuca", "Festuca", "Trifolium", "Festuca"),
  cover   = c(30, 50, 10, 20)
)
```

`Festuca` in `p1` appears at 30 and 50. `"max"` keeps 0.50, `"sum"` adds the
covers to 0.80, and `"first"` keeps the row order's first value, 0.30.

```{r dup-rules}
c(
  max   = as.matrix(specvec(dd, "plot", "species", abundance = "cover",
                            duplicates = "max")$COV)["p1", "Festuca"],
  sum   = as.matrix(specvec(dd, "plot", "species", abundance = "cover",
                            duplicates = "sum")$COV)["p1", "Festuca"],
  first = as.matrix(specvec(dd, "plot", "species", abundance = "cover",
                            duplicates = "first")$COV)["p1", "Festuca"]
)
```

Which rule fits depends on what the duplicate means. `"sum"` reads two rows as
two parts of one plant's cover and adds them; `"max"` and `"first"` read them as
competing estimates and keep one. The count of duplicated pairs is recorded on
the object regardless of the rule, so you can see whether duplicates were present
at all:

```{r dup-meta}
specvec(dd, "plot", "species", abundance = "cover")$meta$n_duplicates
```

`duplicates = "error"` stops with a message naming how many duplicated pairs it
found, instead of aggregating. That is the strict choice for a dataset where a
duplicate signals an upstream problem, not a real repeat measurement: the build
fails loudly so you fix the source, with no quiet averaging over a bug.

## Time and labels

Two more columns ride along when present. `time` is a plot-level time column,
decade or year, stored as a named vector keyed by plot id. `labels` is a
character vector of plot-level column names carried as a data frame, one row per
plot. Both are read once per plot, taken from the first record of each plot, so
they belong to the plot rather than to any single species in it.

```{r time-labels}
dt <- data.frame(
  plot    = c("p1", "p1", "p2", "p2"),
  species = c("Festuca", "Bromus", "Festuca", "Carex"),
  cover   = c(40, 20, 60, 30),
  decade  = c(1980, 1980, 2010, 2010),
  habitat = c("grassland", "grassland", "fen", "fen")
)
xt <- specvec(dt, "plot", "species", abundance = "cover",
              time = "decade", labels = c("habitat"))
xt$time
xt$labels
```

A stored `time` column powers the windowed embeddings and the species and
community trajectories, where the data are split into time slices and a species'
vector is tracked across them; `vignette("specvec-temporal")` runs those. Labels
are carried for the benchmark, where the EUNIS habitat metric scores how well the
embedding separates known habitat classes in the species space;
`vignette("specvec-benchmark")` uses them. The core embedding reads neither
column, so a dataset without time or habitat information still builds and embeds,
and you add these columns only for the analyses that use them.

## Validation and missing data

Before building anything, `specvec()` drops rows with a missing or empty plot id
or species id, treating both `NA` and the empty string `""` as missing. A record
with no taxon name carries no co-occurrence and a record with no plot id cannot
be placed in a site, so neither contributes and both leave the count. If every
row turns out invalid, the build stops rather than returning an empty object.

```{r valid-drop}
dn <- data.frame(
  plot    = c("p1", "p1", "p2", NA),
  species = c("Festuca", "Trifolium", "Festuca", "Carex"),
  cover   = c(40, 10, 60, 80)
)
specvec(dn, "plot", "species", abundance = "cover")$meta$n_obs
```

The table has 4 rows; one has an `NA` plot, so `n_obs` reports the 3 valid
records that survived. Reading `n_obs` against the row count of the source table
is the input check this vignette keeps returning to: a number lower than expected
means rows were dropped, and the gap tells you how many. An unrecognized
Braun-Blanquet code becomes `NA` with a warning that names the offending codes,
so a typo surfaces instead of silently mapping to a cover.

```{r valid-bb-warn}
cover_from_scale(c("r", "2", "xx"), scale = "braun_blanquet")
```

Missing cover values, whether absent in the source or produced by an
unrecognized code, are filled with the median of the observed covers and then
clamped to `[0, 1]`, so `COV` has no holes for the engine to trip on. The median
is a neutral stand-in that does not pull the cover distribution toward an extreme,
and the warning still fires, so an imputed value is visible rather than silent.

## On your own data

A continental table such as the European Vegetation Archive (EVA) maps straight
onto the same call. EVA is an access-controlled database, available on request
rather than as an open download, and specvec bundles none of it. The archive
exports one row per plot-species observation already, so the work is naming
columns: point each role at its column and read the cover
on the scale the archive uses. `data.table::fread()` handles a multi-million-row
export at a speed base R reading cannot match, and the sparse matrices keep the
built object small even when the species count runs into the tens of thousands.

```{r eva, eval = FALSE}
library(data.table)
eva <- fread("eva_export.csv")
x <- specvec(
  eva,
  plot      = "PlotObservationID",
  species   = "taxon",
  abundance = "cover",
  time      = "Decade",
  labels    = c("EUNIS"),
  cover_scale = "percent"
)
emb <- species_embedding(x, method = "abund_pmi", dim = 64, min_occurrence = 5)
```

From here the object feeds `species_embedding()` and the rest of the pipeline.
`vignette("specvec-methods")` defines what the embedding computes.
